---
title: "Allometric equations for mangrove biomass estimation: Evaluating multiple variables in mixed effects models using a global data compilation"
author: Benjamin Branoff & Charles Price.
date: April 8, 2026
output: github_document
---

# Description
This is the repository for the journal article:
Price et al. (2026) Local vs global allometric equations: Evaluating multiple variables in mixed effects models for biomass estimation using a global mangrove data compilation. For the more general 'treellometry' package, visit the ['master' branch](https://github.com/BBranoff/treellometry/master).


# Abstract
Accurate estimation of mangrove biomass is key component of the blue carbon economy and allometric equations are a cornerstone of methodological approaches from plot to landscape scales. A perennial question facing investigators is whether global allometric equations can be applied to their biome of interest, or whether local trees need to be harvested to provide accurate estimates. Further, which combinations of tree dimensions provide the most accurate estimates is not always clear. We examined 16 different allometric equations for four tree measures within a global data set for three prominent trans-Atlantic mangroves species. We used mixed effects models with site location and tree species modeled as random effects, and tree DBH, height, canopy diameter and wood density as fixed effects. While the mixed effects model with the greatest number of parameters was the best performing model based on five different statistical measures with a mean error of 1.392 kg, the improvement over simpler models was marginal. Simply measuring the DBH of each tree would only add an additional 380 grams of error per tree compared to the best performing model. These results provide qualified support for the use of global equations and can help subsequent investigators to prioritize data collection efforts.

# Methods
The methods below can be done manually, by downloading the referenced data and functions from this repository and sourcing them in R, or they can be done by installing the package in R as shown here. 

```{r false-load-data,results='hide',eval=FALSE,echo=TRUE}
##  to install this repository directly, the installation function (remotes) first need to be installed
install.packages("remotes")
##  then install from github
remotes::install_github("BBranoff/treellometry@ecoinf")
library(treellometry)
library(dplyr)
```
```{r load-data,results='hide',echo=FALSE,message=FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
library(sf)
library(patchwork)
devtools::load_all()
options(width = 1000)
#saveRDS(list(mods=mods,mods_scaled=mods_scaled,mods5cm=mods5cm,mods5cm_scaled=mods5cm_scaled,
#             mods_sameN =mods_reducedN, mods_sameN5cm =mods5cm_reducedN),"E:\\OneDrive - USDA\\Chuck ME models\\R Package\\mods.rds")
mods <- readRDS("E:/OneDrive - USDA/Chuck ME models/R Package/mods.rds")
mods5cm <- mods$mods5cm
mods_reducedN <- mods$mods_sameN
mods5cm_reducedN <- mods$mods_sameN5cm
mods <- mods$mods


data("mangroves")
###  set the variables of interest
responsevar <- "AGB.kg"
predictorvars <- c("DBH.cm","Height.m","CanopyDiameter.m","WoodDensity.g.cm3","Comp.DBH.H.Den","Comp.Can.H.Den")
dat <- mangroves |>
  mutate(Comp.DBH.H.Den = (DBH.cm^2)*Height.m*WoodDensity.g.cm3,
         Comp.Can.H.Den = (CanopyDiameter.m^2)*Height.m*WoodDensity.g.cm3)
dat[[paste0("log", responsevar)]] <- log(dat[[responsevar]])
for (var in predictorvars) {
  dat[[paste0("log", var)]] <- log(dat[[var]])
}
###  the random effect variables
groupvars = c("Species", "Site")
```
```{r Fig1,results='hide',echo=FALSE,message=FALSE,warning=FALSE,fig.height=5, message=FALSE,warning=FALSE,fig.cap="Fig. 1 Plot locations for the 10 studies considered in our analysis. Location symbol color and the color bar correspondscorrespond the maximum single tree ABG (kg) at each site."}
ggplot(mangroves |> 
         group_by(Site)|>
         summarise(AGB.kg=max(AGB.kg,na.rm=TRUE),Long=mean(Long),Lat=mean(Lat))|>
         mutate(Long=if_else(Site=="San Juan",Long+2,Long))|>
         st_as_sf(coords=c("Long","Lat"),crs=4326))+
  geom_sf(data=rnaturalearth::ne_coastline())+
  geom_sf(pch=21,size=3,aes(fill=AGB.kg))+
  scale_fill_viridis_b()+
  geom_sf_text(aes(label =Site),position=position_jitter(height=4))+
  ylim(-25,40)+xlim(-95,10)+
  theme_bw()+
  theme(axis.title = element_blank())
```

With the library successfully loaded, the following methods will demonstrate the workflow documented in the above publication. 
The first steps are to create the composite variables (product of variables). These are meant to roughly represent tree form (volume) and are commonly used in allometric modelling because two allometric covariables are likely to be correlated, which destabilizes estimates in  multivariate models. Creating a composite variable eliminates this issue, although it does introduce interpretation considerations, as the product of variables isn’t always directly relatable to tree form.
Here, we also set the response variable, which in our case is always aboveground biomass, as well as the various predictor variables. We also establish the grouping variables, species and location. These will be used to model random effects, which aims to investigate a central premise of this publication, that species and location are significant and substantial in influencing biomass predictions from allometric measurements. In short, mixed effects models will 'partition' the variance in biomass associated with individual species and locations and then attempt to model the remaining variance based on the allometric predictor variables. This is very useful in determining how much of the biomass is species or location specific, and how much depends primarily on the allometric measurements. If a significant and substantial portion of the variance in biomass can be attributed to these grouping variables, that may justify the harvest of local trees to determine location and species specific allometric models. On the other hand, if the mixed effects models determine that an "insubstantial" portion of the variance is due to these grouping variables, harvesting local trees is probably not warranted and the use of 'global' models that apply to all species and all locations is 'good enough'. Ultimately, that decision lies with local managers and scientists, but here we outline a statistical workflow to help make that informed decision. 
For a more detailed outline of mixed-effects modelling in ecology, see [Harrison et al. (2018)](https://doi.org/10.7717/peerj.4794).

Then, we run an important transformation on the data, a log transformation of all response and predictor variables, which is very common in allometry (see [Gingerich (2000)](https://doi.org/10.1006/jtbi.2000.2008), and [Kerkhoff and Enquist (2009)](https://doi.org/10.1016/j.jtbi.2008.12.026)). It is important to note that transformations fundamentally change model coefficients and their interpretation of the original data. So it is important to note here that coefficients and performance metrics for this publication are reported on models from the log transformed data. Thus, a slope and intercept term reported from this workflow will represent the log of variable X and the log of variable Y, and predictions will represent the log of variable Y (biomass, in our case). In a later section we demonstrate how to back-transform predictions to provide a linear (non-log) representation of biomass and the Root Mean Square Error (RMSE).     

```{r prep-data,eval=FALSE,warning=FALSE,message=FALSE}
data("mangroves")

###  create composite variables
mangroves <- mangroves |>
  mutate(Comp.DBH.H.Den = (DBH.cm^2)*Height.m*WoodDensity.g.cm3,
         Comp.Can.H.Den = (CanopyDiameter.m^2)*Height.m*WoodDensity.g.cm3)

###  set the variables of interest
responsevar <- "AGB.kg"
predictorvars <- c("DBH.cm","Height.m","CanopyDiameter.m","WoodDensity.g.cm3","Comp.DBH.H.Den","Comp.Can.H.Den")
###  the random effect variables
groupvars = c("Species", "Site")
##  we want to perform models on log-transformed data
mangroves[[paste0("log", responsevar)]] <- log(mangroves[[responsevar]])
for (var in predictorvars) {
    mangroves[[paste0("log", var)]] <- log(mangroves[[var]])
}
responsevar <- paste0("log",responsevar)
predictorvars <- paste0("log",predictorvars)

mods <- explore_allom_models(mangroves,responsevar,predictorvars,groupvars)
mods_reducedN <- explore_allom_models(mangroves,responsevar,predictorvars,groupvars,SameN=TRUE)
###  5cm or greater
mods5cm <- explore_allom_models(mangroves|>filter(DBH.cm>=5),responsevar,predictorvars,groupvars)
mods5cm_reducedN <- explore_allom_models(mangroves|>filter(DBH.cm>=5),responsevar,predictorvars,groupvars,SameN=TRUE)
```

As the models run, we can first examine the above discussed variance. Below is a general visual of the variable distributions. These 'kernel density estimates' or distributions are a representation of the probability density of any given value within the data. Tall and skinny distributions have relatively little variance compared to short and flat distributions. These visualizations help understand how the data are distributed among the groups and how the individual groups fit into the overall 'population', which is an underlying basis for testing in mixed effects models.  

Notice how the data is much more spread out among sites than species. This suggests site level variability in mangrove tree allometry is more than that at the species level. Also notice that all variables are on the same general scale ([-3,3]), which means we dont necessarily have to scale them to the same order of magnitude. This would be important for mixed effects models utilizing predictor variables from very different orders of magnitude.

```{r pdf,fig.height=8, echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Fig.2 Distributions of the predictor variables. Notice the greater variance between locations, compared to that between species. Wood density is not shown as it was species specific and thus only composed of three"}
(ggplot(mangroves |> tidyr::pivot_longer(cols=c(AGB.kg,DBH.cm,Height.m,CanopyDiameter.m))|>
         mutate(value=log10(value),
                Site=factor(Site,levels=c("Biscayne","Benin","San Juan","Bertioga","Everglades","Louisiana","Guadeloupe","Guaratiba","French Guiana","Puerto Rico"))))+
         geom_density(aes(x=value,fill=Site),alpha=0.5)+
  scale_fill_manual(values=c("#B283BA","#FFE76D","#FE8F89","#95D8F5","#66AAD7","#A2B4FE","#E89875","#C67282","#ADCD82","#F4D078"))+
  facet_wrap(~name)+
  theme_classic())/
(ggplot(mangroves |> tidyr::pivot_longer(cols=c(AGB.kg,DBH.cm,Height.m,CanopyDiameter.m))|>
         mutate(value=log10(value)))+
         geom_density(aes(x=value,fill=Species),alpha=0.5)+
  facet_wrap(~name)+
  theme_classic())
      
```


## A look inside 'explore_allom_models()'
The above use of the 'explore_allom_models()' accomplishes the bulk of the analysis for this publication, which includes fitting models of biomass from various combinations of predictor variables and fixed and mixed effects, as well as retrieving model coefficients and fitness metrics used in model evaluation. Rather than leaving all of that in the black box of the function, here we break down the internal steps of 'explore_allom_models()' to provide more insight.

First, we begin to loop through each of the predictor variables and build and fit models with those variables. Here, we use k to denote the number of predictor variables in the equation. A k of 1 is a simple regression and a k greater than one is multiple regression with more than one predictor. Two important exclusions are happening below. The first is that the 'WoodDensity' variable is not included in simple regression because its values are generalized for each species and are not location specific, thus it does not meaningfully influence biomass for individual trees. Second, is that composite variables are not included in multiple regression because they are already a combination of multiple variables and it simply doesnt make sense in our case to include the influence of a variable twice in the same model.   
```{r explore_allom_models2, eval=FALSE}
for (k in 1:length(predictorVars)) {
    if (k==1){
      ### create predictor combination sets
      ### exclude wood density because it is only from the San Juan dataset and only relative to entire species, not individuals
      combos <- combn(predictorVars[!grepl("WoodDensity",predictorVars)], k, simplify = FALSE)
    }else{
      ####  exclude the composite variables because they are already combined, doesnt make sense to include them in multivariate models
      if (grepl("Comp",predictorVars[k])){next}
      combos <- combn(predictorVars[!grepl("Comp",predictorVars)], k, simplify = FALSE)
    }
}
```

The 'combos' variable created in the above portion of the script is a set of variable combinations that will be used in the model. For this publication, variable importance was not evaluated, thus we only consider one permutation of a given combination of variables. However, we do perform various combinations of fixed and mixed effects for each permutation. 

Next, each predictor set is used in both a fixed effects model as well as various mixed effects models. Running the below script produces one fixed effects model for biomass from the combination of predictor variables. This is repeated for each set of predictors and each model is added to a list of all the models, with the name reflecting the 'family' of the model (basically, which set of predictors was used) as well as the level of fixed or mixed effects. These names are important for identifying and evaluating model families in post-hoc analyses. 
```{r fixed-mod,eval=FALSE}
preds <- combos
##  get the log-transformed version of the variable
log_preds <- paste0("log", preds)
###  remove missing values
data_clean <- dat |> filter(across(c(log_preds,paste0("log", responseVar)), ~ !is.na(.)))
###  create the model formulas, fixed effects first
fixed_formula <- reformulate(log_preds, response = paste0("log", responseVar))
###  run the model
###  first fixed effects
fixed_model <- lm(fixed_formula, data = data_clean)
##   add the model to the list and name it appropriately
mods <- append(mods,list(fixed_model))
names(mods)[length(mods)] <- paste0(model,".Fixed")
###  get summary stats
fixed_summary <- summary(fixed_model)
##  run cross validation and get the mse for each fold
cv <- cv::cv(fixed_model, data_clean, k=10, details=TRUE)
##  Akaike and Bayesian Information Criteria
AIC_fixed <- AIC(fixed_model)
BIC_fixed <- BIC(fixed_model)
##  Root Mean Squared Error
RMSE_fixed <- sqrt(mean(resid(fixed_model)^2))
##  The standardized RMSE, as a fraction of the mean response variable
#RMSE_fixed_std <-  RMSE_fixed/mean(data_clean[,paste0("log", responseVar)],na.rm=TRUE)
##  The cross validated mean of the RMSE
RMSE_CVmean_fixed <-  mean(sqrt(cv$details$criterion),na.rm=TRUE)
##  The cross validated standard deviation of the RMSE
RMSE_CVsd_fixed <-  sd(sqrt(cv$details$criterion),na.rm=TRUE)
## r-squared
R2_fixed <- fixed_summary$r.squared
##  sigma
sig_fixed <-  sigma(fixed_model)
##  variance inflation factor, only valid on multiple regression models
if (k>1){VIF_fixed <- car::vif(fixed_model)}else{VIF_fixed <- NA}
###  predict on the model
predsF <- predsF |> left_join(data.frame(ID=data_clean$ID,pred=predict(fixed_model,newdata=data_clean))|>
                                ###  add in the backtransformed prediction using the Baskerville method
                                ###  this is useful for calculating non-skewed errors in post-hoc analyses
                                mutate(pred_btrans = 10^(pred+((summary(fixed_model)$sigma^2)/2))),by=join_by(ID))
###  get the coefficient names
nmsf <- c(names(coef(fixed_model)),names(coef(fixed_model))[-1])
###  make the coefficent names more generic so they can be added to a table with other model coefficients
nmsf[2:(2*length(preds)+1)] <- c(paste0("slope.var",1:(length(preds))),paste0("VIF.var",1:(length(preds))))
nmsm <- nmsf
###  create a data frame row to hold the coefficients
###  this will be added to a growing dataframe of model coefficients
fixed_cols <- as_tibble_row(setNames(as.list(c(coef(fixed_model),VIF_fixed)),
                                     paste0(nmsf, "_Fixed")))
```
The result of the above is one row of a dataframe with some standard information about the model. As an example, below is the result for a two-variable model of biomass as a function of DBH and height. The table includes the coefficients for the intercept and the slopes attributed to each of the variables, as well as the variance inflation factor (VIF) associated with each predictor. Notice here that although VIF is not really a coefficient and is more of a metric of model performance and validity, we include it in the coefficients table because there is a value associated with each variable, much like each slope coefficient. Notice also that we identify each variable as var1, var2, etc., instead of their actual names, because we will include them all in the same table and need to maintain consistent column names. The performance metrics of the fixed effects models are demonstrated further below, as they will be included with those from mixed effects models from the same set of predictor variables.    
```{r fixed-mod-example,echo=FALSE,warnings=FALSE}
fixed_model <- mods$mods[['6.Fixed']]
fixed_summary <- summary(fixed_model)
###  remove missing values
data_clean <- dat |> filter(if_all(c(logHeight.m,logDBH.cm,logAGB.kg), ~ !is.na(.)))
cv <- cv::cv(fixed_model, data_clean, k=10, details=TRUE)
AIC_fixed <- AIC(fixed_model)
BIC_fixed <- BIC(fixed_model)
RMSE_fixed <- sqrt(mean(resid(fixed_model)^2))
#RMSE_fixed_std <-  RMSE_fixed/mean(data_clean[,paste0("log", responsevar)],na.rm=TRUE)
RMSE_CVmean_fixed <-  mean(sqrt(cv$details$criterion),na.rm=TRUE)
RMSE_CVsd_fixed <-  sd(sqrt(cv$details$criterion),na.rm=TRUE)
R2_fixed <- fixed_summary$r.squared
sig_fixed <-  sigma(fixed_model)
VIF_fixed <- car::vif(fixed_model)
nmsf <- c(names(coef(fixed_model)),names(coef(fixed_model))[-1])
nmsf[2:length(nmsf)] <- c(paste0("slope.var",1:2),paste0("VIF.var",1:2))
fixed_cols <-as_tibble_row(setNames(as.list(c(coef(fixed_model),VIF_fixed)),
                                     paste0(nmsf, "_Fixed")))
fixed_cols
```

Next are the more complicated mixed effects models. For every fixed effects model there are six mixed effects models, three for random intercepts only (one for each of the grouping variables and another that has both grouping variables) and three also for random intercepts and slopes. In the below code, these model structures are created but the models are not yet run.    
```{r mixed-mod,eval=FALSE}
###  mixed effects
###  random intercept, with all grouping variables
mixed_formula_int <- as.formula(paste0(
  deparse(fixed_formula,width.cutoff = 100L),
  " + ",
  paste0("(1|", groupVars, ")", collapse = " + "))
)
###  random intercept and slope, with all grouping variables
mixed_formula_int_slope <- as.formula(paste0(
  deparse(fixed_formula,width.cutoff = 100L),
  " + ",
  paste0("(",log_preds,"|", rep(groupVars,length(preds)), ")", collapse = " + "))
)
###  now add in random effects on individual grouping variables
for (m in 1:length(groupVars)){
  mixed_formula_int <- append(mixed_formula_int,
                              as.formula(paste0(
                                deparse(fixed_formula,width.cutoff = 100L),
                                " + ",
                                "(1|", groupVars[m], ")")))
names(mixed_formula_int)[length(mixed_formula_int)] <- paste0("MixedInt_",groupVars[m])
mixed_formula_int_slope <- append(mixed_formula_int_slope,
                                  as.formula(paste0(
                                    deparse(fixed_formula,width.cutoff = 100L),
                                    " + ",
                                    paste0("(",log_preds,"|", rep(groupVars[m],length(preds)), ")", collapse = " + "))))
names(mixed_formula_int_slope)[length(mixed_formula_int_slope)] <- paste0("MixedIntSlope_",groupVars[m])
}
names(mixed_formula_int)[1] <- paste0("MixedInt_",paste(groupVars,collapse="&"))
names(mixed_formula_int_slope)[1] <- paste0("MixedIntSlope_",paste(groupVars,collapse="&"))
###  create the random effects combination options
mmmods <- c(paste0(paste(groupvars,collapse=""),c("int","intslope")),outer(groupvars,c("int","intslope"),paste0))
```

As an example of the above product, below are all of the model structures for the first set of models, which use DBH as a predictor. The first model, the fixed effects model has already been run at this point, but the remaining models have not, we have only established their structure. We will run each one individually in the next step, as well as collect coefficient and performance information. 
```{r print-mods,echo=FALSE}
forms = do.call(rbind,lapply(1:7, function(x){
  data.frame(model= deparse1(formula(mods$mods[[x]])),
             name=names(mods$mods[x]))
  }))
forms$description <- c("fixed effects model", "both species and location as random intercepts", "both species and location as random intercepts and slopes", "species as random intercepts", "species as random intercepts and slopes","location as random intercepts", "location as random intercepts and slopes")
print(forms)
```

In the penultimate step, we iterate through the above created mixed effects model structures and attempt to fit them. We use a 'tryCatch' statement to avoid stopping the routine in the case of a model that, for whatever reason, throws an error. This is more common in mixed effects models due to their complex structure and requirements. If the fit is successful, we extract the desired information, including the same metrics we gathered for the fixed effects version, as well as the Intraclass Correlation Coefficient, which is important in evaluating mixed effects models [Snijders & Bosker, 2012](https://www.stats.ox.ac.uk/~snijders/mlbook.htm). Most of the other metrics are equivalent to the fixed effects metrics, however the r-squared returned here is the 'conditional r-squared'. This is the r-squared value considering both fixed and mixed effects. See MuMIn::r.squaredGLMM for more information. In the end, we add together the collected metrics and coefficients for the fixed effects model and all of the mixed effects models for the current predictor set.

```{r ME-mods-noeval,eval=FALSE}
###  for each of the ME model formula templates, run the combinations
for(mm in 1:length(mixed_formula_int)){
  MMods <- c(mixed_formula_int[mm],mixed_formula_int_slope[mm])
  if (mm==1){ME = paste(groupVars,collapse="&")}else{ME = groupVars[mm-1]}
  ##  Fit mixed model
  ##  create lists to hold coefficients and model metrics
  R2_mixed <- sigs_mixed <- AIC_mixed <-BIC_mixed<-RMSE_mixed<-coefs_mixed <- ICC <- VIF_mixed <- list()
  for (M in 1:length(MMods)){
    ###  fit the model, if possible given the data
    mixed_model <- tryCatch({
      lmer(MMods[M][[1]], data = data_clean, REML = FALSE)
      }, error = function(e) NULL)
    ###  if the model ran successfully, get the coefficients and metrics
    if (!is.null(mixed_model)) {
      ###  first the predictions
      redsMM <- predsMM |> left_join(data.frame(ID=data_clean$ID,pred=predict(mixed_model,newdata=data_clean))|>
                                       ###  add in the backtransformed prediction using the Baskerville method
                                       ###  this is useful for calculating non-skewed errors in post-hoc analyses
                                                mutate(pred_btrans = 10^(pred+((summary(mixed_model)$sigma^2)/2))),by=join_by(ID))
      ##  add the model to the list and name it appropriately
      mods <- append(mods,list(mixed_model))
      names(mods)[length(mods)] <- paste(model,names(MMods[M]),sep=".")
      # r-squared, in this case we are using the 'conditional r-squared', which includes both fixed and random effects 
      R2_mixed <- append(R2_mixed,MuMIn::r.squaredGLMM(mixed_model)[, "R2c"])
      # sigma
      sigs_mixed <- append(sigs_mixed,sigma(mixed_model))
      # Aikaike and Bayes Information Criteria
      AIC_mixed <- append(AIC_mixed,AIC(mixed_model))
      BIC_mixed <- append(BIC_mixed,BIC(mixed_model))
      ## Root mean squared error
      RMSE_mixed <- append(RMSE_mixed ,sqrt(mean(resid(mixed_model)^2)))
      ## The standardized RMSE, again against the mean of the response variable
      # RMSE_mixed_std = append(RMSE_mixed_std,sqrt(mean(resid(mixed_model)^2))/mean(data_clean[, paste0("log", responseVar)],na.rm=TRUE))
      ## The mean of the cross validated RMSE
      RMSE_CVmean_mixed <- append(RMSE_CVmean_mixed , mean(sqrt(cv$details$criterion),na.rm=TRUE))
      ## The standard deviation of the cross validated RMSE
      RMSE_CVsd_mixed <- append(RMSE_CVsd_mixed , sd(sqrt(cv$details$criterion),na.rm=TRUE))
      #  variance inflation factor
      if (k>1){VIF_mixed <- car::vif(mixed_model)}else{VIF_mixed <-NA}
      ##  add the coefficients together
      coefs_mixed <- append(coefs_mixed,c(fixef(mixed_model,add.dropped=TRUE),VIF_mixed))
      #  get the intraclass correlation coefficient
      ICC <- append(ICC,icc(mixed_model)[1])
      } else {
      ####  if the model could not run, set its metrics and coefficients to NA
      predsMM <- cbind(predsMM,rep(NA,nrow(predsMM)))
      R2_mixed <- append(R2_mixed,NA)
      sigs_mixed <- append(sigs_mixed,NA)
      AIC_mixed <- append(AIC_mixed,NA)
      BIC_mixed <- append(BIC_mixed,NA)
      RMSE_mixed <- append(RMSE_mixed,NA)
      #RMSE_mixed_std <- append(RMSE_mixed_std,NA)
      RMSE_CVmean_mixed <- append(RMSE_CVmean_mixed,NA)
      RMSE_CVsd_mixed <- append(RMSE_CVsd_mixed,NA)
      ICC <- append(ICC,NA)
      VIF_mixed <- append(VIF_mixed,NA)
      coefs_mixed[[M]] <- setNames(rep(NA, length(log_preds) + 1),
                                   c("(Intercept)", log_preds))
      }
  }
  ### with both the fixed and mixed effects models fit for the same set of predictors, we add their collected information to a dataset where
  ##  they can be more easily compared. 
  result_row <- tibble(
    VarGroup = varGroup,
    Model = paste(preds, collapse = ", "),
    NumPredictors = length(preds),
    Rsq_Fixed = R2_fixed,
    Sig_Fixed = sig_fixed,
    AIC_Fixed = AIC_fixed,
    BIC_Fixed = BIC_fixed,
    RMSE_Fixed = RMSE_fixed,
    #RMSE_Fixed_Std = RMSE_fixed_std,
    RMSE_CVmean_Fixed = RMSE_CVmean_fixed,
    RMSE_CVsd_Fixed = RMSE_CVsd_fixed,
    ) |> mutate(MixedEffects=ME) |>
    bind_cols(as_tibble_row(setNames(as.list(c(R2_mixed,sigs_mixed,AIC_mixed,BIC_mixed,RMSE_mixed,RMSE_mixed_std,RMSE_CVmean_mixed,RMSE_CVsd_mixed,ICC)),
                                     paste0(rep(c("Rsq_Mixed","Sig_Mixed","AIC_Mixed","BIC_Mixed","RMSE_Mixed","RMSE_Mixed_Std","RMSE_CVmean_Mixed","RMSE_CVsd_Mixed","ICC_Mixed"),each=2),
                                            c("Int","IntSlope")))))
  coef_row <- tibble(
    VarGroup = varGroup,
    MixedEffects=ME,
    Model = paste(preds, collapse = ", "),
    NumPredictors = length(preds)) |>
    bind_cols(fixed_cols)|>
    bind_cols(as_tibble_row(setNames(as.list(coefs_mixed),
                                     paste0(nmsm, "_Mixed",rep(c("Int","IntSlope"),each=length(nmsm))))))
  results[[length(results) + 1]] <- result_row
  coefs[[length(coefs) + 1]] <- coef_row
}

```

The result of the above is a dataframe containing relevant metrics and model coefficents for all of the models within a given 'family', that is all of the models that use the same set of predictors but with different combinations of mixed effects. This is done for all of the families, and the results are combined together into a master data frame containing all of the information from all of the models. Notice how the first half of the columns in the below tables are all the same because they relate to the fixed effects only. Values begin to change in the right-most sides of the tables, which reflect differences in mixed-effects. These tables are designed to be able to compare the fixed effects values with the different mixed-effects values, thus the fixed-effects information is repeated for each row to allow for easier comparisons across each row rather than down each column (although that can still be done to compare the mixed effects models).  

```{r ME-mods,echo=FALSE,message=FALSE,warning=FALSE}
results <- list()
coefs <- list()
MMods <- mods$mods[grep("7.M",names(mods$mods))]
MMods <- MMods[1:6]
ME <- sapply(strsplit(names(MMods),"_"),"[[",2)
for (me in unique(ME)){
  MMods2 <- MMods[ME==me]
  R2_mixed <- sigs_mixed <- AIC_mixed <-BIC_mixed<-RMSE_mixed<-coefs_mixed <- ICC <- VIF_mixed <- list()
  for (M in 1:length(MMods2)){
    mixed_model <-MMods2[[M]]
    R2_mixed <- append(R2_mixed,MuMIn::r.squaredGLMM(mixed_model)[, "R2c"])
    sigs_mixed <- append(sigs_mixed,sigma(mixed_model))
    AIC_mixed <- append(AIC_mixed,AIC(mixed_model))
    BIC_mixed <- append(BIC_mixed,BIC(mixed_model))
    RMSE_mixed <- append(RMSE_mixed ,sqrt(mean(resid(mixed_model)^2)))
    VIF_mixed <- car::vif(mixed_model)
    coefs_mixed <- append(coefs_mixed,c(lme4::fixef(mixed_model,add.dropped=TRUE),VIF_mixed))
    ICC <- append(ICC,performance::icc(mixed_model)[1])
  }
  result_row <- tibble(
  VarGroup = 6,
  Model = paste(c("Height.m","DBH.cm"), collapse = ", "),
  NumPredictors = 2,
  Rsq_Fixed = R2_fixed,
  Sig_Fixed = sig_fixed,
  AIC_Fixed = AIC_fixed,
  BIC_Fixed = BIC_fixed,
  RMSE_Fixed = RMSE_fixed,
  ) |> mutate(MixedEffects=me) |>
  bind_cols(as_tibble_row(setNames(as.list(c(R2_mixed,sigs_mixed,AIC_mixed,BIC_mixed,RMSE_mixed,ICC)),
                                 paste0(rep(c("Rsq_Mixed","Sig_Mixed","AIC_Mixed","BIC_Mixed","RMSE_Mixed","ICC_Mixed"),each=2),
                                          c("Int","IntSlope")))))
  coef_row <- tibble(
  VarGroup = 6,
  MixedEffects=ME[M],
  Model =paste(c("Height.m","DBH.cm"), collapse = ", "),
  NumPredictors = 2) |>
  bind_cols(fixed_cols)|>
  bind_cols(as_tibble_row(setNames(as.list(coefs_mixed),
                                    paste0(nmsf, "_Mixed",rep(c("Int","IntSlope"),each=length(nmsf))))))
  results[[length(results) + 1]] <- result_row
  coefs[[length(coefs) + 1]] <- coef_row
}
do.call(rbind,results)
do.call(rbind,coefs)
```

We have now collected all relevant models, with varying combinations of predictor variables and mixed effects. The 'mods' object contains the actual models as well as tables of their coefficients, various model performance metrics, and predictions from the models. 
We will now sort through, filter, and rank the models based on these performance metrics. Focusing first on the model performance metrics, we first arrange and rank the models based on their performance.Here we also take the important step of calculating a back-transformed RMSE value for each model. It has been shown that taking a simple exponential of a log transformed value will reflect the geometric mean, not the average, which is intended [Baskerville, 1971](https://cdnsciencepub.com/doi/epdf/10.1139/x72-009). Therefore, we apply a correction factor to the predicted values to back-transform them to the linear scale, and then we calculate the residuals and the linear RMSE from those values. This correction takes the form of: $$estimate = exp(\hat{y}+\frac{\sigma^2}{2})$$
where $\hat{y}$ is the log scaled prediction from the model and $\sigma^2$ is the square of the sigma value returned by the model, essentially the residual standard error of the model. 

We also calculate the Root Median Square Error (RMdSE). This metric helps evaluate the per-tree error rather than the per-dataset error that is more represented by the RMSE. Because large trees are exceptionally rare, and because an ordinary least squares regression (such as the ones we are performing) will be best representative of the median tree sizes, error terms for the outliers will typically be much larger and these errors will strongly skew the RMSE towards the greater values. While the RMSE is still a good metric to compare models for full-dataset error, the RMdSE is better suited to reflect how much error is 'typical' for each tree. 

```{r results,message=FALSE,warning=FALSE}
##  for model comparison, we only use the models that were all fit to the same data
##  this same process is repeated for the >5cm dbh dataset as well 
##  take the results and round the values for aesthetics 
performance <- mods_reducedN$results |>
  mutate(across(7:33,~round(.x,5)))
###  get the back transformation of the RMSE
###  to do this, need to back transform the predictions and recalculate the RMSE
RMSE_lin <- lapply(seq_along(mods_reducedN$mods), function(x) {
  Model=names(mods_reducedN$mods)[x]
  x <- mods_reducedN$mods[[x]]
  # 1. Safe class checking and R2 extraction
  if (!inherits(x, "lmerMod")) { 
    sigma_val <- summary(x)$sigma
  } else { 
    # Extracts the residual standard deviation for mixed models
    sigma_val <- sigma(x) 
  }
  # Extract actual response vector from the model frame
  resp_log <- model.response(model.frame(x))
  # back transform the response to linear scale
  resp <- exp(resp_log)
  # Generate predictions safely on the log scale
  preds_log <- predict(x) 
  # Calculate Log-scale RMSE consistently
  RMSE <- sqrt(mean((resp_log - preds_log)^2, na.rm = TRUE))
  # Apply Baskerville correction and back-transform
  preds_lin <- exp(preds_log) * exp((sigma_val^2) / 2)
  # Calculate back-transformed RMSEn and RMdSE
  RMSE_lin <- sqrt(mean((resp - preds_lin)^2, na.rm = TRUE))
  RMdSE_lin <- sqrt(median((resp - preds_lin)^2, na.rm = TRUE))
  # calculate relative RMSE using the mean AGB for the model
  meanAGB <- mean(resp ,na.rm=TRUE)
  medianAGB <- median(resp ,na.rm=TRUE)
  # Return data frame
  data.frame(Model=Model,RMSE = RMSE, RMSE_lin = RMSE_lin,RMdSE_lin=RMdSE_lin, meanAGB=meanAGB,medianAGB=medianAGB,RelRMSE_lin=100*RMSE_lin/meanAGB,
             RelRMdSE_lin=100*RMdSE_lin/medianAGB)
})
RMSE_lin <- do.call(rbind,RMSE_lin)
##  pivot the table and categorize metrics and models
performance_long <- performance |>
  tidyr::pivot_longer(Rsq_Fixed:ICC2_MixedIntSlope,names_sep = "_",names_to =c("Metric","Effects")) |>
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) |>
  filter(Metric!="ICC2") |>
  distinct(across(1:8),.keep_all = TRUE)|>
  mutate(Metric=if_else(grepl("RMSE",Metric),gsub("RMSE","RMSE_log",Metric),Metric))
##  add in the linear RMSE values
performance_long <- bind_rows(performance_long,
                          performance_long %>% select(VarGroup,Model,ModelN,NumPredictors,NumObservations,MixedEffects,Effects,ModelName) %>% distinct()|>
  left_join(RMSE_lin %>% select(-c(meanAGB)) %>%tidyr::pivot_longer(cols=-Model,names_to="Metric") |> filter(Metric %in% c("RMdSE_lin","RelRMdSE_lin","RMSE_lin","RelRMSE_lin")),by=join_by(ModelName==Model)))

##  return back to wide format
performance_wide <- performance_long |> select(-ModelName)|>tidyr::pivot_wider(names_from=c("Metric",Effects))|>group_by(ModelN)|>
  tidyr::fill(everything(),.direction = "down")|>filter(!is.na(MixedEffects))|>
  relocate(RMdSE_lin_Fixed,RelRMdSE_lin_Fixed,RMSE_lin_Fixed,RelRMSE_lin_Fixed,.before=RMSE_log_Fixed)|>
  relocate(RMdSE_lin_MixedInt,RelRMdSE_lin_MixedInt,RMSE_lin_MixedInt,RelRMSE_lin_MixedInt,.before=RMSE_log_MixedInt)|>
  relocate(RMdSE_lin_MixedIntSlope,RelRMdSE_lin_MixedIntSlope,RMSE_lin_MixedIntSlope,RelRMSE_lin_MixedIntSlope,.before=RMSE_log_MixedIntSlope)

##  rank the models based on their metrics
performance_ranked <- performance_long
performance_ranked <- rbind(performance_ranked%>% filter(Effects!="Fixed"),
                        ##  we separate the fixed effects models from the mixed effects models because the fixed effects
                        ##  information is repeated in the rows pertaining to the same model family.
                        ##  So, we can then remove the duplicated fixed effects information for brevity. 
                        performance_ranked %>% filter(Effects=="Fixed")|>
                          distinct(Model,Metric,.keep_all = TRUE)|>
                          mutate(MixedEffects=NA)) |>
  ###  for each metric, rank the values from each model.
  group_by(Metric) |>
  ###  first across all models 
  ###  If R-squared or ICC, we want the highest value but all other values are optimized at the minimum
  mutate(globalrank=if_else(Metric %in% c("Rsq","ICC"),rank(-value),rank(value))) |>
  mutate(globalrank=if_else(Metric =="ICC"&(is.na(value)|value<0.1),NA,globalrank)) |>
  ungroup() |>
  group_by(Metric,VarGroup) |>
  ###  then the same within model "families"
  mutate(familyrank=if_else(Metric %in% c("Rsq","ICC"),rank(-value),rank(value))) |>
  mutate(familyrank=if_else(Metric =="ICC"&(is.na(value)|value<0.1),NA,familyrank)) |>
  ungroup() |>
  ##  now pivot back to wider with the rankings
  tidyr::pivot_wider(names_from =c("Metric"),values_from = c("value","globalrank","familyrank"))  |>
  ## remove singular ranking
  select(-contains("rank_Singular")) |>
  rowwise() |>
  ##  compute the mean of the ranks for each model family, this we call the 'mean rank'
  ##  here, we dont rank models with an ICC less than 0.1 or a missing ICC value
  ##  These suggest random effects are either minimal (zero) or there is not enough information to compute
  mutate(globalranks_mean=mean(c(globalrank_Rsq,globalrank_AIC,globalrank_BIC,globalrank_RMSE_lin,globalrank_RMdSE_lin,globalrank_Sig,globalrank_RMSE_log.CVsd)),
         globalranks_mean=if_else((is.na(value_ICC)|value_ICC<.1)&Effects!="Fixed",NA, globalranks_mean),
         familyranks_mean=mean(c(familyrank_Rsq,familyrank_AIC,familyrank_BIC,familyrank_RMSE_lin,familyrank_RMdSE_lin,familyrank_Sig,familyrank_RMSE_log.CVsd)),
         familyranks_mean=if_else((is.na(value_ICC)|value_ICC<.1)&Effects!="Fixed",NA, familyranks_mean)) |>
  ungroup() |>
  ###  now create a master global rank based on the means
  ###  next we will create a master master, and so on...
  mutate(masterrank_global=rank(globalranks_mean))  |>
  group_by(VarGroup) |>
  ##  same for the family ranks
  mutate(masterrank_family=rank(familyranks_mean)) |>
  ungroup() |>
  relocate(masterrank_global,masterrank_family,.after=Effects) |>
  arrange(masterrank_global)

##  attach the VIF values from the coefficients table
##  this time, coefficients come from the full dataset models
coef <- mods$coefs |>
  mutate(across(6:32,~round(.x,5)))|>
  rowwise()|>
  mutate(maxVIF_Fixed=max(c_across(c(VIF.var1_Fixed,VIF.var2_Fixed,VIF.var3_Fixed,VIF.var4_Fixed)),na.rm=TRUE),
         maxVIF_MixedInt=max(c_across(c(VIF.var1_MixedInt,VIF.var2_MixedInt,VIF.var3_MixedInt,VIF.var4_MixedInt)),na.rm=TRUE),
         maxVIF_MixedIntSlope=max(c_across(c(VIF.var1_MixedIntSlope,VIF.var2_MixedIntSlope,VIF.var3_MixedIntSlope,VIF.var4_MixedIntSlope)),na.rm=TRUE))
coef_long <- coef %>%
  tidyr::pivot_longer('(Intercept)_Fixed':maxVIF_MixedIntSlope,names_sep = "_",names_to =c("Coefficient","Effects")) %>%
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) %>%
  distinct(across(1:7),.keep_all = TRUE)
coef_wide <- coef_long %>%
  tidyr::pivot_wider(names_from=Coefficient)%>%
  relocate(ModelN,Effects,.before=MixedEffects)

performance_ranked <- performance_ranked %>% 
  left_join(coef_wide%>%select(ModelN,MixedEffects,Effects,maxVIF),by=join_by(ModelN==ModelN,MixedEffects==MixedEffects,Effects==Effects))%>%
  relocate(maxVIF,.after=value_ICC)|>
  relocate(value_RMdSE_lin,value_RelRMdSE_lin,value_RMSE_lin,value_RelRMSE_lin,.before=value_RMSE_log)

coefs_filtered <- coef_wide %>%
  filter(paste0(ModelN,MixedEffects,Effects) %in% paste0(performance_ranked$ModelN,performance_ranked$MixedEffects,performance_ranked$Effects)) %>%  relocate(ModelN)
coefs_ranked <- coefs_filtered[match(paste0(performance_ranked$ModelN,performance_ranked$MixedEffects,performance_ranked$Effects),
                                       paste0(coefs_filtered$ModelN,coefs_filtered$MixedEffects,coefs_filtered$Effects)), ]
performance_coeffs_combined <- performance_ranked %>%select(ModelN,Model,NumObservations,MixedEffects,Effects,masterrank_global,value_Rsq:value_RMSE_log,value_RMSE_lin,value_RelRMSE_lin,value_RMdSE_lin,value_RelRMdSE_lin,value_RMSE_log.CVmean:value_ICC) %>%
  left_join(coef_wide%>%select(-c(VarGroup,Model,NumPredictors)),by=join_by(ModelN==ModelN,MixedEffects==MixedEffects,Effects==Effects))
```
```{r save-tables,echo=FALSE,message=FALSE,warning=FALSE}
performance5cm <- mods5cm_reducedN$results%>%
  mutate(across(7:33,~round(.x,5)))
RMSE5cm_lin <- lapply(seq_along(mods5cm_reducedN$mods), function(x) {
  Model=names(mods5cm_reducedN$mods)[x]
  x <- mods5cm_reducedN$mods[[x]]
  # 1. Safe class checking and R2 extraction
  if (!inherits(x, "lmerMod")) { 
    sigma_val <- summary(x)$sigma
  } else { 
    # Extracts the residual standard deviation for mixed models
    sigma_val <- sigma(x) 
  }
  # Extract actual response vector from the model frame
  resp_log <- model.response(model.frame(x))
  # back transform the response to linear scale
  resp <- exp(resp_log)
  # Generate predictions safely on the log scale
  preds_log <- predict(x) 
  # Calculate Log-scale RMSE consistently
  RMSE <- sqrt(mean((resp_log - preds_log)^2, na.rm = TRUE))
  # Apply Baskerville correction and back-transform
  preds_lin <- exp(preds_log) * exp((sigma_val^2) / 2)
  # Calculate back-transformed RMSE
  RMSE_lin <- sqrt(mean((resp - preds_lin)^2, na.rm = TRUE))
  RMdSE_lin <- sqrt(median((resp - preds_lin)^2, na.rm = TRUE))
  # calculate relative RMSE using the mean AGB for the model
  meanAGB <- mean(resp ,na.rm=TRUE)
  medianAGB <- median(resp ,na.rm=TRUE)
  # Return data frame
  data.frame(Model=Model, RMSE_lin = RMSE_lin, RMdSE_lin= RMdSE_lin, meanAGB=meanAGB,medianAGB=medianAGB,RelRMSE_lin=100*RMSE_lin/meanAGB,
             RelRMdSE_lin=100*RMdSE_lin/medianAGB)
})
RMSE5cm_lin <- do.call(rbind,RMSE5cm_lin)
RMSE5cm_wide <- RMSE5cm_lin|>
    mutate(ModelN=sapply(strsplit(Model,"\\."),"[[",1),
           MixedEffects=sapply(strsplit(Model,"\\."),"[[",2),
           Effects=if_else(grepl("Effects",MixedEffects),NA,sapply(strsplit(MixedEffects,"_"),"[[",1)))|>rowwise()|>mutate(MixedEffects=gsub(paste0(Effects,"_"),"",MixedEffects))

performance5cm_long <- performance5cm %>%
  tidyr::pivot_longer(Rsq_Fixed:ICC2_MixedIntSlope,names_sep = "_",names_to =c("Metric","Effects")) %>%
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) %>%
  filter(Metric!="ICC2")%>%
  distinct(across(1:8),.keep_all = TRUE)|>
  mutate(Metric=if_else(grepl("RMSE",Metric),gsub("RMSE","RMSE_log",Metric),Metric))
##  add in the linear RMSE values
performance5cm_long <- bind_rows(performance5cm_long,
                          performance5cm_long %>% select(VarGroup,Model,ModelN,NumPredictors,NumObservations,MixedEffects,Effects,ModelName) %>% distinct()%>%
  left_join(RMSE5cm_lin %>% select(-meanAGB) %>%tidyr::pivot_longer(cols=-Model,names_to="Metric")%>% filter(Metric %in% c("RMdSE_lin","RelRMdSE_lin","RMSE_lin","RelRMSE_lin")),by=join_by(ModelName==Model)))

##  return back to wide format
performance5cm_wide <- performance5cm_long |> select(-ModelName)|>tidyr::pivot_wider(names_from=c("Metric",Effects))|>group_by(ModelN)|>tidyr::fill(everything(),.direction = "down")|>filter(!is.na(MixedEffects))|>
  relocate(RMdSE_lin_Fixed,RelRMdSE_lin_Fixed,RMSE_lin_Fixed,RelRMSE_lin_Fixed,.before=RMSE_log_Fixed)|>
  relocate(RMdSE_lin_MixedInt,RelRMdSE_lin_MixedInt,RMSE_lin_MixedInt,RelRMSE_lin_MixedInt,.before=RMSE_log_MixedInt)|>
  relocate(RMdSE_lin_MixedIntSlope,RelRMdSE_lin_MixedIntSlope,RMSE_lin_MixedIntSlope,RelRMSE_lin_MixedIntSlope,.before=RMSE_log_MixedIntSlope)

performance5cm_ranked <- performance5cm_long
performance5cm_ranked <- rbind(performance5cm_ranked%>% filter(Effects!="Fixed"),
                        performance5cm_ranked %>% filter(Effects=="Fixed")%>%
                          distinct(Model,Metric,.keep_all = TRUE)%>%
                          mutate(MixedEffects=NA))%>%
  group_by(Metric)%>%
  mutate(globalrank=if_else(Metric %in% c("Rsq","ICC"),rank(-value),rank(value)))%>%
  mutate(globalrank=if_else(Metric =="ICC"&(is.na(value)|value<0.1),NA,globalrank)) |>
  ungroup()%>%
  group_by(Metric,VarGroup)%>%
  mutate(familyrank=if_else(Metric %in% c("Rsq","ICC"),rank(-value),rank(value)))%>%
  mutate(familyrank=if_else(Metric =="ICC"&(is.na(value)|value<0.1),NA,familyrank)) |>
  ungroup()%>%
  tidyr::pivot_wider(names_from =c("Metric"),values_from = c("value","globalrank","familyrank")) %>%
  ## remove singular ranking
  select(-contains("rank_Singular"))%>%
  rowwise()%>%
  mutate(globalranks_mean=mean(c(globalrank_Rsq,globalrank_AIC,globalrank_BIC,globalrank_RMSE_lin,globalrank_RMdSE_lin,globalrank_Sig,globalrank_RMSE_log.CVsd)),
         globalranks_mean=if_else((is.na(value_ICC)|value_ICC<.1)&Effects!="Fixed",NA, globalranks_mean),
         familyranks_mean=mean(c(familyrank_Rsq,familyrank_AIC,familyrank_BIC,familyrank_RMSE_lin,familyrank_RMdSE_lin,familyrank_Sig,globalrank_RMSE_log.CVsd)),
         familyranks_mean=if_else((is.na(value_ICC)|value_ICC<.1)&Effects!="Fixed",NA, familyranks_mean))%>%
  ungroup()%>%
  mutate(masterrank_global=rank(globalranks_mean)) %>%
  group_by(VarGroup)%>%
  mutate(masterrank_family=rank(familyranks_mean))%>%
  ungroup()%>%
  relocate(masterrank_global,masterrank_family,.after=Effects)%>%
  arrange(masterrank_global)

coef5cm <- mods5cm$coefs |>
  mutate(across(6:32,~round(.x,5)))|>
  rowwise()|>
  mutate(maxVIF_Fixed=max(c_across(c(VIF.var1_Fixed,VIF.var2_Fixed,VIF.var3_Fixed,VIF.var4_Fixed)),na.rm=TRUE),
         maxVIF_MixedInt=max(c_across(c(VIF.var1_MixedInt,VIF.var2_MixedInt,VIF.var3_MixedInt,VIF.var4_MixedInt)),na.rm=TRUE),
         maxVIF_MixedIntSlope=max(c_across(c(VIF.var1_MixedIntSlope,VIF.var2_MixedIntSlope,VIF.var3_MixedIntSlope,VIF.var4_MixedIntSlope)),na.rm=TRUE))
coef5cm_long <- coef5cm %>%
  tidyr::pivot_longer('(Intercept)_Fixed':maxVIF_MixedIntSlope,names_sep = "_",names_to =c("Coefficient","Effects")) %>%
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) %>%
  distinct(across(1:7),.keep_all = TRUE)
coef5cm_wide <- coef5cm_long %>%
  tidyr::pivot_wider(names_from=Coefficient)%>%
  relocate(ModelN,Effects,.before=MixedEffects)

performance5cm_ranked <- performance5cm_ranked %>% 
  left_join(coef5cm_wide%>%select(ModelN,MixedEffects,Effects,maxVIF),by=join_by(ModelN==ModelN,MixedEffects==MixedEffects,Effects==Effects))%>%
  relocate(maxVIF,.after=value_ICC)|>
 relocate(value_RMdSE_lin,value_RelRMdSE_lin,value_RMSE_lin,value_RelRMSE_lin,.before=value_RMSE_log)

coefs5cm_ranked <- coef5cm_wide %>%
  filter(paste0(ModelN,MixedEffects,Effects) %in% paste0(performance5cm_ranked$ModelN,performance5cm_ranked$MixedEffects,performance5cm_ranked$Effects)) %>%  relocate(ModelN)
coefs5cm_ranked <- coefs5cm_ranked[match(paste0(performance5cm_ranked$ModelN,performance5cm_ranked$MixedEffects,performance5cm_ranked$Effects),
                                       paste0(coefs5cm_ranked$ModelN,coefs5cm_ranked$MixedEffects,coefs5cm_ranked$Effects)), ]
performance5cm_coeffs_combined <- performance5cm_ranked %>%select(ModelN,Model,NumObservations,MixedEffects,Effects,masterrank_global,value_Rsq:value_RMSE_log,value_RMSE_lin,value_RelRMSE_lin,value_RMdSE_lin,value_RelRMdSE_lin,value_RMSE_log.CVmean:value_ICC) %>%
  left_join(coef5cm_wide%>%select(-c(VarGroup,Model,NumPredictors)),by=join_by(ModelN==ModelN,MixedEffects==MixedEffects,Effects==Effects))


write.csv(rbind(performance_wide|>mutate(size="all"),
                performance5cm_wide|>mutate(size="5cm"))|>relocate(size),"Tables/PriceBranoff_Performance_all.csv",row.names=F)
write.csv(rbind(coef|>mutate(size="all"),
                coef5cm|>mutate(size="5cm"))|>relocate(size),"Tables/PriceBranoff_coefs_all.csv",row.names=F)
write.csv(rbind(performance_coeffs_combined|>mutate(size="all"),
                performance5cm_coeffs_combined|>mutate(size="5cm"))|>relocate(size),
          "Tables/PriceBranoff_Performance_Coeffs_combined_ranked.csv",row.names=F)
write.csv(rbind(performance_ranked|>mutate(size="all"),
                performance5cm_ranked|>mutate(size="5cm"))|>relocate(size),"Tables/PriceBranoff_Performance_ranked.csv",row.names=F)
write.csv(rbind(coefs_ranked|>mutate(size="all"),
                coefs5cm_ranked|>mutate(size="5cm"))|>relocate(size),"Tables/PriceBranoff_coefs_ranked.csv",row.names=F)
```
The above created tables are stored in 'Tables' folder of this repository.

We now have a dataset (performance_ranked) with the top performing model from each family of predictor variables, ranked by their performance metrics. Remember, however, that these rankings are somewhat arbitrary and based primarily upon our ability to logistically sort a large number of models. Different rankings could be applied for different objectives (prioritize RMSE, for example). Still, this gives us a general idea of which models are performing 'better' than others. We are left with around 100 models, which is still a considerable number. 

We now want to test the assumptions of each model. Again, we are looking for a reasonable way to evaluate a large number of models. Using the convenient 'check_model()' function from the performance package, we can produce excellent visualizations for pertinent assumptions. These are stored in the 'Assumptions' folder of the repository. Upon examining the graphics, it seems many of the top models do a good to decent job of meeting the primary regression model assumptions. Again, because these assumptions have varying consequences depending on the objectives of the model (predictions versus comparisons) it is the user's responsibility to scrutinize these assumption tests according to their objectives. The results of the below are included the "Assumptions" folder and below is one example from the 'top' model.
```{r model-assumptions,eval=FALSE,message=FALSE,warning=FALSE}
######  testing assumptions on the top ranked models
checks <- lapply(unique(performance_ranked$ModelName),function(x){
  pdf(file = paste0("./Assumptions/",x,"-model_checks.pdf"), width = 9, height = 11)
  print(performance::check_model(mods$mods[[x]]))
  dev.off()
})
```

```{r, echo=FALSE, out.width="100%",warning=FALSE,message=FALSE,fig.cap="An example of the assumptions plots for model '20.MixedInt_Species&Site'. Each panel is a visual representation of the model assumptions. Many of the top-performing models seem to be satisfactory in meeting these assumptions, but some are not. All top model assumption plots are stored in the 'Assumptions' folder of the repository."}
tfile <- tempfile(fileext = ".png")
suppressWarnings(png <- pdftools::pdf_convert("Assumptions/20.MixedInt_Species&Site-model_checks.pdf",format="png",filenames=tfile,verbose=FALSE))
knitr::include_graphics(tfile)
```

We can also now begin to examine some of the results more closely. One way to scrutinize the performance of multiple models is through the compare_perfomance function from the performance package. This function will evaluate multiple performance metrics, in our case sigma, RMSE, R2, AIC and BIC, across all of the models and produce spider plots with relative values for all. The spider plots indicate the 'best' model as the one with the largest area, allowing for easier visual comparisons. 

```{r performance1,message=FALSE,results='hide',warning=FALSE,fig.cap="Fig. S1_ An example of the product of the compare_performance() function from the performance package. The size of the polygons indicate performance, with bigger being better. This is a quick visual way to compare multiple models for common performance metrics. In the next step, the chart is emulated across multiple families of models."}
# ####  model peformance charts
# ####  select the top filtered models
 modskeep_df <- performance_ranked %>%
   select(ModelN)
 modskeep <- mods_reducedN$mods[grep(paste0(modskeep_df%>%pull(ModelN),".",collapse="|"),names(mods_reducedN$mods))]
 ###  use an lapply function to iterate over the models and compute the performance comparisons in groups
 perfs <- lapply(modskeep_df%>%pull(ModelN)|>unique(),
                 ##  the performance comparison
                 FUN=function(x,mds){ 
                   perf <- performance::compare_performance(mds[which(sapply(strsplit(names(mods$mods),"\\."),"[[",1)==x)],metrics=c("SIGMA","RMSE","AIC","BIC")) %>%mutate(mod=x)
                 ##  add in progress bar for sanity check
                   cat(paste0("\r",round(100*which(x==unique(modskeep_df$ModelN))/length(unique(modskeep_df$ModelN)))))
                   perf
                 }
                 ,mds=mods_reducedN$mods)
 ###  these can be used individually to assess models within families
# plot(perfs[[1]])
```

However, it may also be pertinent to compare models both within their family as well as among the other families as well. The performance package does not offer this functionality, but we can replicate it. In the below code, we arrange the different metrics among the models in such a way so as to replicate the output of compare_performance(). We then scale the metric values so that they are comparable across all of the families. Although doing this makes it slightly more difficult to compare models within families (because the differences are now relatively small) it allows for direct comparisons with the other families. 

In the plots, the most complex model, with random slopes and intercepts on species and site, seems to perform 'better'. But this is debatable depending on which metrics are prioritized. If BIC (which penalized complexity) is prioritized, a simpler model with only random effects on site, may be optimal. Either way, its clear that in general, for the same set of variable, mixed effects models out-perform the fixed effects versions. This is the primary intention of mixed-effects modelling when hierarchical data is suspected. By partitioning the variance into groups and fitting coefficients separately, mixed-effects models are better able to capture the global and group specific trends.   

```{r performance2,fig.height=10, message=FALSE,warning=FALSE,fig.cap="Fig. S1 Performance charts for all model families. As in Fig. 2, larger polygons indicate better relative performance. The indicated RMSE represents the lowest value within that panel. Because all metric values are log-scaled across the entire model set, differences within families are relatively small but those between the 'best' (top left) and 'worst' (bottom right) models are more apparent."}
###  get the stored results for those models
perfs2 <- lapply(modskeep_df |>
                   pull(ModelN)|>unique(),
                 FUN=function(x) performance_long%>%filter(ModelN==x) |>
                   select(Model,ModelN,ModelName,MixedEffects,Effects,Metric,value) |>
                   #tidyr::pivot_longer(cols=Rsq_Fixed:Sig_MixedIntSlope,names_sep ="_",names_to=c("metric","effects")) |>
                   mutate(Name=paste0("mod",as.character(ModelN),"_",MixedEffects,Effects)) |>
                   tidyr::pivot_wider(names_from="Metric") |>
                   filter(!(Effects=="Fixed"&MixedEffects %in% c("Species","Site")))%>%rename(Sigma=Sig))

perfs2 <- do.call(rbind,perfs2) |>
  mutate(Model.x=if_else(grepl("Mixed",Effects),"lmerMod","lm"),
         Model.y=gsub("log|.cm|.m|.g.cm3","",Model)) |>
  # separate the random effects and the random variables
  # these will help in visualizing among multiple families by using the same colors for each group
  tidyr::separate(Name,sep="(?=int|Int)",remove=FALSE, into=c("RandomVars","RandomEffects")) |>
  tidyr::separate(RandomVars,sep="(?<=_)",remove=FALSE, into=c("Name2","RandomVars"))%>%
  group_by(Name2) |>
  mutate(RandomVars=if_else(Model.x=="lm","none",paste0("Random ",gsub("Mixed","",RandomVars)," on ")),
         RandomVars = if_else(RandomVars=="Random Species&Site on ","Random Species & Site on ",RandomVars),
         RandomEffects=if_else(Model.x=="lm","none",gsub("Int","Int.",RandomEffects)),
         RandomEffects = if_else(RandomEffects=="Int.Slope", "Int. & Slope",RandomEffects),
         ###  create the title for each panel, which includes the minimum RMSE for each family
         Model.y = paste0(Model.y,"\n RMSE: ",round(min(RMSE_lin,na.rm=TRUE)))) |>
  ungroup() |>
  arrange(RMSE_lin) |>
  mutate(Model.y=factor(Model.y,levels=unique(Model.y))) %>%
  tidyr::pivot_longer(cols=c("Rsq","Sigma","AIC","BIC","RMdSE_lin","RMSE_lin","RMSE_log.CVsd"))|>
  group_by(name) |>
  ###  re-scale the metrics so that the spiderwebs are comparable across families
  mutate(value2=if_else(name %in% c("Rsq"),scales::rescale(value,to=c(0.1,1)),scales::rescale(-value,to=c(0.1,1))),
         group=paste(RandomVars,RandomEffects,sep=" "),
         group=if_else(group=="none none","Fixed",group)) |>
  ungroup() %>%
  mutate(name=if_else(name=="RMdSE_lin","RMdSE_\nlin",if_else(name=="RMSE_log.CVsd","RMSE_\nCVsd",if_else(name=="RMSE_lin","RMSE_\nlin",name))),
         name=factor(name,levels=c("AIC","BIC","RMdSE_\nlin","RMSE_\nlin","RMSE_\nCVsd","Rsq","Sigma")))
  
###  plot the performance plots
plot_index=0
perfs2 %>%
  ggplot(aes(x=name,y=value2,color=group,group=group))+
  facet_wrap(~Model.y,scales="free",ncol =4,labeller=label_wrap_gen(width = 20))+
  geom_polygon(linewidth=1,alpha=0)+
  scale_x_discrete(
    labels = function(x) {
      plot_index <<- plot_index + 1
      # Only return labels for the 2nd facet (e.g., cyl == 6)
      if (plot_index == 1) {
        return(x)
      } else {
        return("") # Empties out the text for other facets
      }
    })+
  see::coord_radar()+
  scale_y_continuous(limits=c(0,1))+
  guides(color=guide_legend("Model Type",ncol=2)) +
  theme_bw()+
  theme(axis.text.y =element_blank(),axis.title = element_blank(), axis.ticks=element_blank(),legend.position = "bottom",legend.direction =
        "horizontal",panel.grid = element_line(color = "grey"),panel.background = element_blank())
```

Another subject of interest involving the performance metrics is how well a model performs given the number of variables it considers. Typically models performs better with more predictors, as each predictor should be able to explain even a small amount of variance, which boosts performance. However, overly complex models are not always better, despite better performance. This is because greater complexity introduces greater potential for error and more difficulty in interpretation, which may not be warranted for only a small increase in predictive performance. The below figure demonstrates that for many performance metrics, there are simpler models that perform nearly as good or better than models with more predictors and more complexity. 

```{r Fig3,fig.height=10, message=FALSE,warning=FALSE,fig.cap="Fig. 3 Bivariate plots of our six summary statistics (Panels A-F, R2, AIC, BIC, Sigma, the linear RMSE, and the standard deviation of the cross validated log scale RMSE.) as a function of the number of predictors. The different modelling types; fixed effects, mixed intercept and mixed intercept and slope, are denoted with pink, green and blue circles, respectively. The symbols for each number of predictors are offset slightly  (shifted on the x-axis) to increase visibility, e.g. all of the symbols on the far left of Panel A have just a single predictor. Note that while models with more predictors tend to perform better (higher R2, lower everything else), many of the models with 1 or 2 predictors also perform reasonably well. Please consult Tables 3, S1-3 for specific model details."}
ggplot(performance_long|>filter(Metric %in% c("AIC","BIC","Rsq","RMSE_lin","RMdSE_lin","Sig","RMSE_log.CVsd"))|>
         mutate(NumPredictors=if_else(Effects=="Fixed",NumPredictors-0.1,if_else(Effects=="MixedIntSlope",NumPredictors+0.1,NumPredictors)),
                Metric = if_else(Metric=="Rsq","R squared",if_else(Metric=="Sig","Sigma",if_else(Metric=="RMSE_log.CVsd","RMSE CV standard deviation",
                                                                                                 if_else(Metric=="RMSE_lin","linear RMSE",
                                                                                                         if_else(Metric=="RMdSE_lin","linear RMdSE",Metric))))),
                Metric=factor(Metric,levels=c("R squared","AIC","BIC","Sigma","linear RMSE","linear RMdSE","RMSE CV standard deviation")),
                Effects=if_else(Effects=="MixedInt","Mixed Intercept",if_else(Effects=="MixedIntSlope","Mixed Intercept & Slope",Effects))),
       aes(x=NumPredictors,y=value,col=Effects))+
  geom_point(width = 0.2, height = 0)+
  facet_wrap(~Metric,ncol=2,scales='free_y')+
  theme_classic()+
  guides(color=guide_legend(title=""))+
  xlab("Number of Predictors")+
  theme(axis.title.y=element_blank(),legend.position = "bottom")
```

We can also examine more closely the random effects of the models. Here, we demonstrate two metrics of random effects: first, the estimated variance assigned to each component of the model (total variance, fixed variables, random variables, etc.), and second, the effect size as an indication of how important specific groups (species and site in our case) are in the overall variability in biomass. 

Starting with the overall estimate of explained variance, we use the 'get_variance()' function from the insight package to decompose variance in biomass and how much of it might be explained by different model components. Ultimately, through modeling we are trying to explain why the biomass of one tree might be different than another. These differences are summarised by their variance and we can try to assign this variability (variance) to different variables. In mixed effects modelling, we are pooling this variance into different components (in our case, species and location), and then fitting models of the fixed effects (DBH, Height, etc.) to the remaining variance. Anything left over or unexplained is referred to as the residual variance. 

The below graphic represents the mean and standard error of the mean variance among the different models. The total variance is that of aboveground biomass before any modelling is done. This is slightly different for each model because some observations need to be removed when certain fixed effects are missing. The fixed effects variance is that which is estimated to be explained by the predictor variables (DBH, Height, etc.), and the total random is that which is estimated to be attributed to the combined random effects (species and location). Because some models have random intercepts only and others have random intercepts and slopes, the random variances are further decomposed into the intercept and slope terms for each predictor variable. In the below graphic, we have aggregated all models for simplicity, so we cant see what var1, var2, var3 etc. are, but variable specific graphics are provided in the supplementary materials for more detailed insight. 

These figures demonstrate that overall explanation of biomass variability is high, around 90% on average, and that the vast majority is explained by fixed effects, with relatively little attributed to random effects. This corroborates the notion that species and location information contribute relatively little to allometric biomass relationships in comparison to the morphological measurements.

```{r varpart,fig.width=10,fig.height=8, message=FALSE,warning=FALSE,fig.cap="Fig. 4 Partitioning the variance in aboveground biomass to fixed effects, random effects, and residuals; means and standard errors for each component among all models. Total estimated explained variance is at least 90% on average, and fixed effects dominate, accounting for the vast majority of variance (Fig 4a).  With fixed effects removed (Fig. 4b), its easier to compare the random effects, which account for no more than roughly 2% of the variance. This corroborates the view that species and location level information provide relatively little to the variability in allometric relationships we are testing for. The same plots for each model family are provided in the supplementary materials."}
library(insight)
### apply the 'get_variance()' funtion to all models and format into a data frame.
var_list <- lapply(seq_along(modskeep),function(x){if(class(modskeep[[x]])[1]=="lmerMod"){as.data.frame(get_variance(modskeep[[x]]))|>tibble::rownames_to_column(var="component")|>mutate(component=paste0(gsub("\\d", "",component),"var",as.numeric(gsub("\\D", "",component))+1),component=gsub("varNA",".var1",component))|>select(-contains("cor."))|>rowwise()|>mutate(var.total=sum(c_across(any_of(c("var.fixed","var.random","var.residual"))),na.rm=TRUE))|>tidyr::pivot_longer(cols=contains("var."))|>mutate(Model=names(modskeep)[x],name=gsub("var.","",name),name=if_else(grepl("slope|intercept",name),paste(component,name,sep="."),if_else(name=="random","mean random", name)),component=if_else(!grepl("slope|intercept",name),NA,component))|>distinct(component,name,.keep_all = TRUE)}})
###  remove the non-mixed effects models
var_list <- Filter(Negate(is.null), var_list)
###  combine all model data frames
var_list <- do.call(rbind,var_list) |>
  left_join(performance_ranked |> select(ModelName,Model),by=join_by(Model==ModelName))|>
  mutate(name=factor(name,levels=c("total","fixed","mean random","residual","dispersion","distribution","Site.var1.intercept","Site.var1.slope","Site.var2.intercept","Site.var2.slope",
                                      "Site.var3.intercept","Site.var3.slope","Site.var4.intercept","Site.var4.slope",
                                      "Species.var1.intercept","Species.var1.slope","Species.var2.intercept","Species.var2.slope",
                                      "Species.var3.intercept","Species.var3.slope","Species.var4.intercept","Species.var4.slope")))
var_list_agg <- var_list |>
  group_by(Model)|>mutate(value=100*value/value[name=="total"]) |> filter(!name %in% c("dispersion","distribution","total"))|>
  mutate(name=gsub("var1.|var2.|var3.|var4.","",name))|>
  group_by(name) |> summarise(var.mean=mean(value,na.rm=TRUE),var.sd=sd(value,na.rm=TRUE),var.sem=var.sd/sqrt(n()))|>
  ungroup() |>
  mutate(name=factor(name,levels=c("fixed","mean random","residual","Site.intercept","Site.slope","Species.intercept","Species.slope")))

var_list_models <- var_list |>
  group_by(Model)|>mutate(value=100*value/value[name=="total"]) |> filter(!name %in% c("dispersion","distribution","total"))|> 
  group_by(Model.y,name) |> summarise(var.mean=mean(value,na.rm=TRUE),var.sd=sd(value,na.rm=TRUE),var.sem=var.sd/sqrt(n()))|>
         mutate(Model.y=gsub("\\.cm|\\.m|\\.g\\.cm3","",Model.y),
                Model.y=gsub("CanopyDiameter","Canopy Diameter",Model.y),
                Model.y=gsub("WoodDensity","Wood Density",Model.y),
                Model.y=gsub("Comp.Can.H.Den","Composite: Canopy Diameter x Height x Wood Density",Model.y),
                Model.y = if_else(Model.y=="DBH, Height, Canopy Diameter, Wood Density","DBH, Height,\nCanopy Diameter, Wood Density",
                                  if_else(Model.y=="Height, Canopy Diameter, Wood Density","Height, Canopy Diameter,\nWood Density",
                                          if_else(Model.y=="Composite: Canopy Diameter x Height x Wood Density","Composite: Canopy Diameter\nx Height x Wood Density",if_else(Model.y=="DBH, Canopy Diameter, Wood Density","DBH, Canopy Diameter,\nWood Density",Model.y)))))|>
         ungroup()|>
         group_by(Model.y)|>
         mutate(mean.resid = mean(var.mean[name=="residual"],na.rm=TRUE))|>
         ungroup()|>
         arrange(mean.resid)|>
         mutate(Model.y=factor(Model.y,levels=unique(Model.y)))
####  plot
####  with the fixed variance
ggplot(var_list_agg, 
       aes(x = name, y = var.mean, fill = name)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymax=var.mean+var.sem,ymin=var.mean-var.sem),width=0.2)+
  scale_fill_manual(values=c("black","#1B9E77","white",RColorBrewer::brewer.pal(8,"YlOrRd")[3:4],RColorBrewer::brewer.pal(9,"BuPu")[7:9]))+
  #facet_wrap(~Model.y) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust=1),axis.title.x=element_blank())+
  guides(fill=guide_legend("Variance Groups"))+
  labs(
    title = "Fig. 4a Variance Decomposition - with fixed effects.",
    x = "Variance Component",
    y = "Estimated Variance Assignment, Percent of Total"
  ) 

### without the fixed and residual variance
ggplot(var_list_agg|>filter(!name %in% c("dispersion","distribution","total","fixed","residual")), 
       aes(x = name, y = var.mean, fill = name)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymax=var.mean+var.sem,ymin=var.mean-var.sem),width=0.2)+
  scale_fill_manual(values=c("#1B9E77",RColorBrewer::brewer.pal(8,"YlOrRd"),RColorBrewer::brewer.pal(9,"BuPu")[2:9]))+
  #facet_wrap(~Model.y) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust=1),axis.title.x=element_blank())+
  #annotate("text",label="Site",x=5,y=10)+
  #annotate("text",label="Species",x=13,y=10)+
  #geom_vline(xintercept=c(1.5,9.5),lty=2)+
  guides(fill=guide_legend("Variance Groups"))+
  labs(
    title = "Fig. 4b Variance Decomposition - fixed effects removed.",
    x = "Variance Component",
    y = "Estimated Variance Assignment, Percent of Total"
  ) 
```
```{r varpart2,message=FALSE,warning=FALSE}
####  with the fixed variance
ggplot(var_list_models, 
       aes(x = name, y = var.mean, fill = name)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymax=var.mean+var.sem,ymin=var.mean-var.sem),width=0.2)+
  scale_fill_manual(values=c("black","#1B9E77","white",RColorBrewer::brewer.pal(8,"YlOrRd"),RColorBrewer::brewer.pal(9,"BuPu")[2:9]))+
 facet_wrap(~Model.y) +
  theme_bw() +
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())+
  guides(fill=guide_legend("Variance Groups"))+
  labs(
    title = "Variance Decomposition - with fixed effects.",
    x = "Variance Component",
    y = "Estimated Variance Assignment, Percent of Total"
  ) 
ggsave("./README_files/figure-gfm/FigS3A.jpeg")

### without the fixed and residual variance
ggplot(var_list_models|>filter(!name %in% c("dispersion","distribution","total","fixed","residual")), 
       aes(x = name, y = var.mean, fill = name)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymax=var.mean+var.sem,ymin=var.mean-var.sem),width=0.2)+
   scale_fill_manual(values=c("#1B9E77",RColorBrewer::brewer.pal(8,"YlOrRd"),RColorBrewer::brewer.pal(9,"BuPu")[2:9]))+
  facet_wrap(~Model.y) +
  theme_bw() +
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())+
  #annotate("text",label="Site",x=5,y=10)+
  #annotate("text",label="Species",x=13,y=10)+
  #geom_vline(xintercept=c(1.5,9.5),lty=2)+
  guides(fill=guide_legend("Variance Groups"))+
  labs(
    title = "Variance Decomposition - fixed effects removed.",
    x = "Variance Component",
    y = "Estimated Variance Assignment, Percent of Total"
  ) 
ggsave("./README_files/figure-gfm/FigS3B.jpeg")
```

We can then further inspect the random effects for specific species and locations. Here, for simplicity we examine only the simple regression (uni-variate) models with the most complex combination of mixed effects (random slopes and intercepts on both variables).The below graphic indicates random effects as a standardized deviation from 0, which is at the center of each panel and marked with a dotted line. Any deviation that is farther from zero than the corresponding error bar (solid horizontal 'wings' for each dot) indicates a significant random effect. For the most part, the random effects for these three models are insignificant, which suggests that site and species are not contributing substantially to the allometric models and global models might be warranted. However, there are exceptions. Height within sites demonstrate the largest effect sizes, suggesting that using height as a predictor of biomass may be location dependent. This is consistent with previous studies that have recognized the latitudinal variability in mangrove height and the relationship between height and biomass [Rovai et al. 2021]("https://doi.org/10.1111/geb.13268"), and is an important consideration when utilizing LiDAR or other remote sensing techniques that rely heavily on height metrics for biomass estimation. Species, on the other hand, was not found to be a significant random effect, suggesting that these allometric relationships are consistent across the three Trans-Atlantic species. Future work will expand to other species and test this relationship. 

```{r randomeffects,fig.height=10, message=FALSE,warning=FALSE,fig.cap="Fig. 5 Random effects for the three univariate models. Deviations from the zero line indicate a potentially significant random effect. Species exhibited non-significant random effects, although there were only three species in this study and all were from the Atlantic geo-region. Location, however, does demonstrate some significant random effects, especially for height. This suggests that height as a predictor of biomass may be location specific."}
library(lme4)
random_effects <- rbind(as.data.frame(ranef(mods$mods[[3]])) %>% mutate(pred="DBH"),
                        as.data.frame(ranef(mods$mods[[10]])) %>% mutate(pred="Height"),
                        as.data.frame(ranef(mods$mods[[17]])) %>% mutate(pred="Canopy Diameter"))%>%#,%
  mutate(grp=factor(grp,levels=c(rev(sort(unique(as.character(grp[grpvar=="Species"])))),rev(sort(unique(as.character(grp[grpvar=="Site"])))))))
# lattice::qqmath(ranef(mods_scaled$mods[[3]], strip =TRUE))
ggplot(random_effects %>% mutate(term=if_else(term=='(Intercept)',"Intercept","Slope"))%>%
         filter(pred!="Wood Density"), aes(y = grp, x = condval)) +
  geom_errorbar(aes(xmin = condval - 2*condsd, xmax = condval + 2*condsd),width=0) +
  geom_point(aes(fill=grpvar),shape=21,col="black",size=3) +
  #scale_shape_manual(values=c(21:23))+
  facet_grid(term~pred)+# Assuming 'se_intercept' is available
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_vline(xintercept=0,lty=2,col="darkgrey")+
  guides(fill=guide_legend("Random Effect\nVariable"))+
  xlab("Conditional Mode")+ylab("Group")+
  theme_bw()
```

To test model overfitting and general applicability to 'new' data, we also perform a cross validation. Some modelling processes incorporate a data splitting into 'training' and 'testing' subsets. The 'training' subset is typically around 70% of the original dataset and is used to fit the models. These models are then 'tested' on the other 30% of the data to see how well they perform on unseen observations. This allows us to examine how well the models perform on new data. Cross validation (CV) takes this a bit further by 'rotating' the training and testing subsets, which helps avoid results that may only be a result of a 'lucky' (or unlucky) training set. 

Here, we incorporate 10 'folds' of the data, or 10 rotations of the training-testing subsets and get RMSE values for each of the 10 predictions on the testing subsets. When the RMSE values from the 10 folds are similar, this indicates the models and their coefficients are consistent at predicting new data. When they deviate, however, this indicates that predictions are unstable and may not predict well on new data. 

The below figure demonstrates the CV results from our models. Generally, the results indicate that for models incorporating fixed effects only or random intercepts, predictions are stable and may be applicable to new data. For models incorporating random slopes, however, CV indicates unstable results and non-applicability to new data. 
```{r CV,fig.height=5, message=FALSE,warning=FALSE}
ggplot(performance_ranked,aes(y=value_RMSE_log.CVmean,x=value_RMSE_log)) +
  geom_point(shape=21,col="white",fill="blue",size=3)+
  geom_abline(slope=1,intercept=0)+
  theme_classic()+
  facet_grid(rows=vars(Effects))+
  xlab("RMSE")+ylab("RMSE CV")
```

```{r predictions,fig.height=5, message=FALSE,warning=FALSE,fig.cap="Fig. X "}
preds <- cbind(mods$predsF[,c("Species","Site","AGB.kg","logAGB.kg","preds_12.Fixed")],
               data.frame(mods$predsMM[,"preds_12.MixedInt_Species&Site"]))|>
  rename(preds_12.MixedInt_SpeciesandSite=mods.predsMM....preds_12.MixedInt_Species.Site..)|>
  tidyr::pivot_longer(cols=c("preds_12.MixedInt_SpeciesandSite","preds_12.Fixed"))

ggplot(preds,aes(y=exp(logAGB.kg),x=exp(value),col=Site,shape=Species)) +
  geom_point()+
  geom_abline(intercept=0,slope=1)+
  facet_wrap(~name)
```

<!-- ##  random effects-->
<!-- varcorrs <- lapply(seq_along(modskeep),function(x){if(class(modskeep[[x]])[1]=="lmerMod"){as.data.frame(VarCorr(modskeep[[x]]))|>mutate(Model=names(modskeep)[x])}})-->
<!-- varcorrs <- do.call(rbind,varcorrs) |>-->
<!--   left_join(performance_rank_filtered |> select(ModelName,Model),by=join_by(Model==ModelName))-->

<!-- ggplot(varcorrs |> group_by(Model.y,grp) |> summarise(vcov.mean=mean(vcov,na.rm=TRUE),vcov.sd=sd(vcov,na.rm=TRUE),vcov.sem=vcov.sd/sqrt(n()))|>-->
<!--          mutate(Model.y=gsub("\\.cm|\\.m|\\.g\\.cm3","",Model.y),-->
<!-- Model.y=gsub("CanopyDiameter","Canopy Diameter",Model.y),-->
<!--                 Model.y=gsub("WoodDensity","Wood Density",Model.y),-->
<!--                 Model.y=gsub("Comp.Can.H.Den","Composite: Canopy Diameter x Height x Wood Density",Model.y))|>-->
<!--          ##  arrange by residual variance-->
<!--          ungroup()|>-->
<!--          group_by(Model.y)|>-->
<!--          mutate(mean.resid = mean(vcov.mean[grp=="Residual"],na.rm=TRUE))|>-->
<!--          ungroup()|>-->
<!--          arrange(mean.resid)|>-->
<!--          mutate(Model.y=factor(Model.y,levels=unique(Model.y))),-->
<!--        aes(x = grp, y = vcov.mean, fill = grp))+-->
<!--   geom_bar(stat = "identity", position = "dodge") +-->
<!--   geom_errorbar(aes(ymax=vcov.mean+vcov.sem,ymin=vcov.mean-vcov.sem),width=0.2)+-->
<!--   facet_wrap(~Model.y) +-->
<!--   theme_minimal() +-->
<!--   theme(axis.text.x=element_blank(),axis.title.x=element_blank())+-->
<!--   scale_fill_manual(values=c("#A50026",RColorBrewer::brewer.pal(4,"YlOrRd"),RColorBrewer::brewer.pal(9,"YlGnBu")[5:8]))+-->
<!--   guides(fill=guide_legend("Variance Groups"))+-->
<!--   labs(title = "Comparison of Variance Components",-->
<!--        y = "Variance Estimate")-->

<!-- ###  Fixed effects-->
<!-- coefs_full <- mods_scaled_full$coefs |>-->
<!--   select(MixedEffects,Model,contains("Rsq"))|>-->
<!--   tidyr::pivot_longer(cols=contains("Rsq"),names_pattern="(.*)\\.(.*)_(.*)",names_to=c("metric","var","effects"))|>-->
<!--   filter(!is.na(value))|>-->
<!--   rowwise()|>-->
<!--   mutate(Model=as.character(Model),-->
<!--          var=if_else(!grepl(",",Model),Model,-->
<!--                      if_else(grepl("1",var),strsplit(Model,",")[[1]][1],-->
<!--                      if_else(grepl("2",var),strsplit(Model,",")[[1]][2],-->
<!--                              if_else(grepl("3",var),strsplit(Model,",")[[1]][3],-->
<!--                                      if_else(grepl("4",var),strsplit(Model,",")[[1]][4],NA))))),-->
<!--          var = gsub(" ","",var))-->
<!-- ggplot(coefs_full) +-->
<!--   geom_boxplot(aes(x=var,y=value))+-->
<!--   facet_grid(vars(MixedEffects),vars(effects))+-->
<!--   theme_bw()+-->
<!--   theme(axis.text.x=element_text(angle=90))-->
<!-- ggplot(coefs_full) +-->
<!--   geom_boxplot(aes(x=var,y=value))+-->
<!--   theme_bw()+-->
<!--  theme(axis.text.x=element_text(angle=90))-->

<!-- 
<!-- predsMM <-mods$predsMM -->
<!-- predsF <-mods$predsF -->
<!-- ####  variation in models at each site? -->
<!-- gg_preds1 <- predsMM %>% -->
<!--   #select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>% -->
<!--   left_join(predsF |>select(ID,contains("preds_")),by=join_by(ID))%>% -->
<!--   tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>% -->
<!--   tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>% -->
<!--   mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars), -->
<!--          RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")), -->
<!--          RandomEffects = gsub("^.*?\\.","",RandomEffects)) %>% -->
<!--   filter(!grepl("Site",RandomVars))%>% -->
<!--   ggplot(aes(x=logDBH.cm,y=logAGB.kg,col=Species))+ -->
<!--   geom_point(alpha=1)+ -->
<!--   geom_line(aes(y=value,lty=RandomEffects))+ -->
<!--   #geom_smooth(method="lm",formula=y~x,aes(lty=RandomEffects),se=FALSE)+ -->
<!--   facet_grid(Species~Site)+ -->
<!--   #scale_color_manual(values=c("black",RColorBrewer::brewer.pal(3,"Set1")))+ -->
<!--   scale_linetype_manual(values=c(1,2,4))+ -->
<!--   theme_bw() -->
<!-- gg_preds2 <- predsMM %>% -->
<!--   #select(Dataset:logWoodDensity,contains("mod1_")) %>% -->
<!--   cbind(predsF %>% select(contains("preds_")))%>%#rename(mod1_none_none = mod1_)) %>% -->
<!--   tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>% -->
<!--   tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>% -->
<!--   mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars), -->
<!--          RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")), -->
<!--          RandomEffects = gsub("^.*?\\.","",RandomEffects)) %>% -->
<!--   ggplot(aes(x=logDBH.cm,y=logAGB.kg,shape=Species))+ -->
<!--   #geom_point(col="darkgrey")+ -->
<!--   #geom_smooth(method="lm",formula=y~x,aes(lty=RandomEffects),se=FALSE)+ -->
<!--   facet_wrap(~Species)+ -->
<!--   scale_color_manual(values=c("black",RColorBrewer::brewer.pal(3,"Set1")))+ -->
<!--   scale_linetype_manual(values=c(1,2,4))+ -->
<!--   geom_line(aes(y=value,lty=RandomEffects,col=RandomVars))+ -->
<!--   theme_bw() -->

<!-- #####   Residual density plot -->
<!-- predsMM %>% -->
<!--   select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>% -->
<!--   cbind(predsF %>% select(contains("preds_")))%>% -->
<!--   tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>% -->
<!--   tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>% -->
<!--   mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars), -->
<!--          RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")), -->
<!--          RandomEffects = gsub("^.*?\\.","",RandomEffects), -->
<!--          Residuals = logAGB.kg-value) %>% -->
<!--   ggplot(aes(x=Residuals,col=RandomVars))+ -->
<!--   geom_density()+ -->
<!--   facet_wrap(~RandomEffects)+ -->
<!--   theme_bw() -->

<!-- ###  Residuals versus observations -->
<!-- predsMM %>% -->
<!--   select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>% -->
<!--   cbind(predsF %>% select(contains("preds_")))%>% -->
<!--   tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>% -->
<!--   tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>% -->
<!--   mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars), -->
<!--          RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")), -->
<!--          RandomEffects = gsub("^.*?\\.","",RandomEffects), -->
<!--          Residuals = logAGB.kg-value) %>% -->
<!--   ggplot(aes(x=value,y=Residuals,col=RandomEffects))+ -->
<!--   geom_point()+ -->
<!--   facet_wrap(~RandomVars)+ -->
<!--   theme_bw() -->


