---
title: "ecoinfromatics publication placeholder"
author: Ben Branoff
date: April 8, 2026
output: github_document
---
# Coming Soon
This is the future site of the repository for the journal article:
Price et al. (2026) Local vs global allometric equations: Evaluating multiple variables in mixed effects models for biomass estimation using a global mangrove data compilation.


# Abstract
Accurate estimation of mangrove biomass is key component of the blue carbon economy and allometric equations are a cornerstone of methodological approaches from plot to landscape scales. A perennial question facing investigators is whether global allometric equations can be applied to their biome of interest, or whether local trees need to be harvested to provide accurate estimates. Further, which combinations of tree dimensions provide the most accurate estimates is not always clear. We examined 16 different allometric equations for four tree measures within a global data set for three prominent trans-Atlantic mangroves species. We used mixed effects models with site location and tree species modeled as random effects, and tree DBH, height, canopy diameter and wood density as fixed effects. While the mixed effects model with the greatest number of parameters was the best performing model based on five different statistical measures with a mean error of 1.392 kg, the improvement over simpler models was marginal. Simply measuring the DBH of each tree would only add an additional 380 grams of error per tree compared to the best performing model. These results provide qualified support for the use of global equations and can help subsequent investigators to prioritize data collection efforts.

# Methods
The below methods can be done manually, by downloading the referenced data and functions from this repository and sourcing them in R, or they can be done by installing the package in R. 

```{r false-load-data,results='hide',eval=FALSE,echo=TRUE}
##  to install this repository directly, the installation function (remotes) first need to be installed
install.packages("remotes")
##  then install from github
remotes::install_github("BBranoff/treellometry@ecoinf")
library(treellometry)
```
```{r load-data,results='hide',echo=FALSE,message=FALSE}
devtools::load_all()
options(width = 1000)
#saveRDS(list(mods=mods,mods_scaled=mods_scaled),"E:\\OneDrive - USDA\\Chuck ME models\\R Package\\mods.rds")
mods <- readRDS("E:/OneDrive - USDA/Chuck ME models/R Package/mods.rds")
mods_scaled <- mods$mods_scaled
mods <- mods$mods
```

With the library successfully loaded, the following methods will demonstrate the workflow documented in the above publication. The first steps are to create the composite variables. These are commonly used in allometric modelling because two allometric covariables are likely to be correlated, violating the assumptions of multivariate models. Creating a composite variable eliminates this issue, although it does introduce interpretation considerations.
Here, we also set the response variable, which in our case is always aboveground biomass, as well as the various predictor variables. We also establish the grouping variables, these will be used to model random effects, which aims to investigate a central premise of this publication, that species and location are significant and substantial in influencing biomass predictions from allometric measurements. In short, mixed effects models will 'partition' the variance in biomass associated with individual species and locations and then attempt to model the remaining variance based on the allometric predictor variables. This is very useful in determining how much of the biomass is species or location specific, and how much depends primarily on the allometric measurements. If a significant and substantial portion of the variance in biomass can be attributed to these grouping variables, that justifies the harvest of local trees in order to determine location and species specific allometric models. On the other hand, if the mixed effects models determine that an "insubstantial" portion of the variance is due to these grouping variables, harvesting local trees is probably not warranted and the use of 'global' models that apply to all species and all locations is 'good enough'. Ultimately, that decision lies with local managers and scientists, but here we provide the statistical information necessary to make that informed decision. 
For a more detailed outline of mixed-effects modelling in ecology, see [Harrison et al. (2018)](https://doi.org/10.7717/peerj.4794). 


```{r prep-data,eval=FALSE,warning=FALSE,message=FALSE}
data("mangroves")

###  create composite variables
mangroves <- mangroves |>
  mutate(Comp.DBH.H.Den = (DBH.cm^2)*Height.m*WoodDensity.g.cm3,
         Comp.Can.H.Den = (CanopyDiameter.m^2)*Height.m*WoodDensity.g.cm3)

###  set the variables of interest
responsevar <- "AGB.kg"
predictorvars <- c("DBH.cm","Height.m","CanopyDiameter.m","WoodDensity.g.cm3","Comp.DBH.H.Den","Comp.Can.H.Den")
###  the random effect variables
groupvars = c("Species", "Site")

mods <- explore_allom_models(mangroves,responsevar,predictorvars,groupvars)
mods_scaled <- explore_allom_models(mangroves,responsevar,predictorvars,groupvars,scle=TRUE)
```


## A look inside 'explore_allom_models()'
The above use of the 'explore_allom_models()' accomplishes the bulk of the analysis for this publication, which includes fitting models of biomass from various combinations of predictor variables and fixed and mixed effects, as well as retrieving model coefficients and fitness metrics used in model evaluation. Rather than leaving all of that in the black box of the function, here we break down the internal steps of 'explore_allom_models()' to provide more insight.

The first step is to check to make sure inputs are valid and then to run two important transformations on the data. The first is a log transformation, which is very common in allometry (see [Gingerich (2000)](https://doi.org/10.1006/jtbi.2000.2008), and [Kerkhoff and Enquist (2009)](https://doi.org/10.1016/j.jtbi.2008.12.026)). The second is a scaling that is important for mixed-effects model performance and interpretability [Harrison et al. (2018)](https://doi.org/10.7717/peerj.4794). It is important to note here, however, that transformations fundamentally change model coefficients and their interpretation of the original data. For this reason, coefficients for this publication are reported on models from the non-scaled data.  

```{r explore_allom_models1, eval=FALSE}
  ##  check inputs
  if (responseVar %in% predictorVars) stop("Response variable found in predictor variables")
  # Log-transform response and predictors
  dat[[paste0("log", responseVar)]] <- log(dat[[responseVar]])
  for (var in predictorVars) {
    dat[[paste0("log", var)]] <- log(dat[[var]])
  }
  if (scle==TRUE){
    dat <- dat |>
      mutate( across(contains(predictorVars),function(x) scale(x)[,1]))#,.names = "{paste0(col, '_scaled')}"))
  }
  dat <- dat |> mutate( ID=1:n())
  predsMM <- predsF <- dat
  results <- list()
  coefs <- list()
  mods <- list()
  varGroup=model=0
```

Next, we begin to loop through each of the predictor variables and build and fit models with different combinations if those variables. Here, we use k to denote the number of predictor variables in the equation. A k of 1 is a simple regression and a k greater than one is multiple regression with more than one predictor. Two important exclusions are happening below. The first is that the 'WoodDensity' variable is not included in simple regression because its values are generalized for each species and are not location specific. Second, is that composite variables are not included in multiple regression because they are already a combination of multiple variables and it simply doesnt make sense in our case to include the influence of a variable twice in the same model.   
```{r explore_allom_models2, eval=FALSE}
for (k in 1:length(predictorVars)) {
    if (k==1){
      ### create predictor combination sets
      ### exclude wood density because it is only from the San Juan dataset and only relative to entire species, not individuals
      combos <- combn(predictorVars[!grepl("WoodDensity",predictorVars)], k, simplify = FALSE)
    }else{
      ####  exclude the composite variables because they are already combined, doesnt make sense to include them in multivariate models
      if (grepl("Comp",predictorVars[k])){next}
      combos <- combn(predictorVars[!grepl("Comp",predictorVars)], k, simplify = FALSE)
    }
```

The 'combos' variable created in the above portion of the script is a set of variable combinations that will be used in the model. They are sent along to the next step, which further expands upon the 'combos' by computing all possible combinations of those variables. This is important because evaluating variable importance depends upon the order in which a variable is present within a multiple regression model. As an example, here is the result of combining all of the predictor variables and then permutating their combinations

```{r combo-vars}
combos <- combn(predictorvars[!grepl("Comp",predictorvars)], 4, simplify = FALSE)
for (predss in combos) {
      ### establish all potential combinations of the current predictor set (change the order of predictors).
      perms <- gtools::permutations(length(predss),length(predss),predss)
      if (nrow(perms)==0){perms <- matrix(1,nrow = 1,ncol=1)}
}
print(perms)
```

Next, each of the permutations in 'perms' (each row of the above table) is used in both a fixed effects model as well as various mixed effects models. The result of the below is one fixed effects model for biomass from the unique combination of predictor variables. This is repeated for each unique combination and each model is added to a list of all the models, with the name reflecting the 'family' of the model (basically, which set of predictors was used) as well as the level of fixed or mixed effects. These names are important for identifying and evaluating model families in post-hoc analyses. 
```{r fixed-mod,eval=FALSE}
###  unseen in this script is that p is one row of the above printed table, basically one set of predictor variables. 
###  The function loops through each of the rows to create different models.
preds <- perms[p,]
##  get the log-transformed version of the variable
log_preds <- paste0("log", preds)
###  remove missing values
data_clean <- dat |> filter(across(c(log_preds,paste0("log", responseVar)), ~ !is.na(.)))
###  create the model formulas, fixed effects first
fixed_formula <- reformulate(log_preds, response = paste0("log", responseVar))
###  run the model
###  first fixed effects
fixed_model <- lm(fixed_formula, data = data_clean)
##   add the model to the list and name it appropriately
mods <- append(mods,list(fixed_model))
names(mods)[length(mods)] <- paste0(model,".Fixed")
###  get summary stats
fixed_summary <- summary(fixed_model)
##  run cross validation and get the mse for each fold
cv <- cv::cv(fixed_model, data_clean, k=10, details=TRUE)
##  Akaike and Bayesian Information Criteria
AIC_fixed <- AIC(fixed_model)
BIC_fixed <- BIC(fixed_model)
##  Root Mean Squared Error
RMSE_fixed <- sqrt(mean(resid(fixed_model)^2))
##  The standardized RMSE, as a fraction of the mean response variable
RMSE_fixed_std <-  RMSE_fixed/mean(data_clean[,paste0("log", responseVar)],na.rm=TRUE)
##  The cross validated mean of the RMSE
RMSE_CVmean_fixed <-  mean(sqrt(cv$details$criterion),na.rm=TRUE)
##  The cross validated standard deviation of the RMSE
RMSE_CVsd_fixed <-  sd(sqrt(cv$details$criterion),na.rm=TRUE)
## r-squared
R2_fixed <- fixed_summary$r.squared
##  sigma
sig_fixed <-  sigma(fixed_model)
##  variance inflation factor, only valid on multiple regression models
if (k>1){VIF_fixed <- car::vif(fixed_model)}else{VIF_fixed <- NA}
###  predict on the model
predsF <- predsF |> left_join(data.frame(ID=data_clean$ID,pred=predict(fixed_model,newdata=data_clean)),by=join_by(ID))
###  get the coefficient names
nmsf <- c(names(coef(fixed_model)),names(coef(fixed_model))[-1])
###  make the coefficent names more generic so they can be added to a table with other model coefficients
nmsf[2:(2*length(preds)+1)] <- c(paste0("slope.var",1:(length(preds))),paste0("VIF.var",1:(length(preds))))
nmsm <- nmsf
###  create a data frame row to hold the coefficients
###  this will be added to a growing dataframe of model coefficients
fixed_cols <- as_tibble_row(setNames(as.list(c(coef(fixed_model),VIF_fixed)),
                                     paste0(nmsf, "_Fixed")))
```
The result of the above is one row of a dataframe with some standard information about the model. As an example, here is the result for a two-variable model of biomass as a function of DBH and height:
```{r fixed-mod-example,echo=FALSE}
fixed_model <- mods$mods[['7.Fixed']]
fixed_summary <- summary(fixed_model)
cv <- cv::cv(fixed_model, data_clean, k=10, details=TRUE)
AIC_fixed <- AIC(fixed_model)
BIC_fixed <- BIC(fixed_model)
RMSE_fixed <- sqrt(mean(resid(fixed_model)^2))
RMSE_fixed_std <-  RMSE_fixed/mean(data_clean[,paste0("log", responseVar)],na.rm=TRUE)
RMSE_CVmean_fixed <-  mean(sqrt(cv$details$criterion),na.rm=TRUE)
RMSE_CVsd_fixed <-  sd(sqrt(cv$details$criterion),na.rm=TRUE)
R2_fixed <- fixed_summary$r.squared
sig_fixed <-  sigma(fixed_model)
VIF_fixed <- car::vif(fixed_model)
nmsf <- c(names(coef(fixed_model)),names(coef(fixed_model))[-1])
nmsf[2:length(nmsf)] <- c(paste0("slope.var",1:2),paste0("VIF.var",1:2))
fixed_cols <-as_tibble_row(setNames(as.list(c(coef(fixed_model),VIF_fixed)),
                                     paste0(nmsf, "_Fixed")))
fixed_cols
```

Next are the more complicated mixed effects models. For every fixed effects model there are six mixed effects model, three for random intercepts only (one for each of the grouping variables and another that has both grouping variables) and three also for random intercepts and slopes. In the below code, these model structures are created but the models are not yet run.    
```{r mixed-mod,eval=FALSE}
###  mixed effects
###  random intercept, with all grouping variables
mixed_formula_int <- as.formula(paste0(
  deparse(fixed_formula,width.cutoff = 100L),
  " + ",
  paste0("(1|", groupVars, ")", collapse = " + "))
)
###  random intercept and slope, with all grouping variables
mixed_formula_int_slope <- as.formula(paste0(
  deparse(fixed_formula,width.cutoff = 100L),
  " + ",
  paste0("(",log_preds,"|", rep(groupVars,length(preds)), ")", collapse = " + "))
)
###  now add in random effects on individual grouping variables
for (m in 1:length(groupVars)){
  mixed_formula_int <- append(mixed_formula_int,
                              as.formula(paste0(
                                deparse(fixed_formula,width.cutoff = 100L),
                                " + ",
                                "(1|", groupVars[m], ")")))
names(mixed_formula_int)[length(mixed_formula_int)] <- paste0("MixedInt_",groupVars[m])
mixed_formula_int_slope <- append(mixed_formula_int_slope,
                                  as.formula(paste0(
                                    deparse(fixed_formula,width.cutoff = 100L),
                                    " + ",
                                    paste0("(",log_preds,"|", rep(groupVars[m],length(preds)), ")", collapse = " + "))))
names(mixed_formula_int_slope)[length(mixed_formula_int_slope)] <- paste0("MixedIntSlope_",groupVars[m])
}
names(mixed_formula_int)[1] <- paste0("MixedInt_",paste(groupVars,collapse="&"))
names(mixed_formula_int_slope)[1] <- paste0("MixedIntSlope_",paste(groupVars,collapse="&"))
###  create the random effects combination options
mmmods <- c(paste0(paste(groupvars,collapse=""),c("int","intslope")),outer(groupvars,c("int","intslope"),paste0))
```

As an example of the above product, below are all of the model structures for the first set of models, which use DBH as a predictor. The firs model, the fixed effects model has already been run at this point, but the remaining models have not, we have only established their structure at this point. We will run each one individually in the next step, as well as collect coefficient and performance information. 
```{r print-mods,echo=FALSE}
forms = do.call(rbind,lapply(1:7, function(x){
  data.frame(model= deparse1(formula(mods$mods[[x]])),
             name=names(mods$mods[x]))
  }))
forms$description <- c("fixed effects model", "both species and location as random intercepts", "both species and location as random intercepts and slopes", "species as random intercepts", "species as random intercepts and slopes","location as random intercepts", "location as random intercepts and slopes")
print(forms)
```

In the penultimate step, we iterate through the above created mixed effects model structures and attempt to fit them. We use a 'tryCatch' statement to avoid stopping the routine in the case of a model that, for whatever reason, throws an error. This is more common in mixed effects models due to their complex structure and requirements. If the fit is successful, we extract the desired information, including the same metrics we gathered for the fixed effects version, as well as the Intraclass Correlation Coefficient, which is important in evaluating mixed effects models [Snijders & Bosker, 2012](https://www.stats.ox.ac.uk/~snijders/mlbook.htm). Most of the other metrics are equivalent to the fixed effects metrics, however the r-squared returned here is the 'conditional r-squared'. This is the r-squared value considering both fixed and mixed effects. See MuMIn::r.squaredGLMM for more information. In the end, we add together the collected metrics and coefficients for the fixed effects model and all of the mixed effects model for the current predictor set.

```{r ME-mods-noeval,eval=FALSE}
###  for each of the ME model formula templates, run the combinations
for(mm in 1:length(mixed_formula_int)){
  MMods <- c(mixed_formula_int[mm],mixed_formula_int_slope[mm])
  if (mm==1){ME = paste(groupVars,collapse="&")}else{ME = groupVars[mm-1]}
  ##  Fit mixed model
  ##  create lists to hold coefficients and model metrics
  R2_mixed <- sigs_mixed <- AIC_mixed <-BIC_mixed<-RMSE_mixed<-coefs_mixed <- ICC <- VIF_mixed <- list()
  for (M in 1:length(MMods)){
    ###  fit the model, if possible given the data
    mixed_model <- tryCatch({
      lmer(MMods[M][[1]], data = data_clean, REML = FALSE)
      }, error = function(e) NULL)
    ###  if the model ran successfully, get the coefficients and metrics
    if (!is.null(mixed_model)) {
      ###  first the predictions
      predsMM <- predsMM |> left_join(data.frame(ID=data_clean$ID,pred=predict(mixed_model,newdata=data_clean)),by=join_by(ID))
      ##  add the model to the list and name it appropriately
      mods <- append(mods,list(mixed_model))
      names(mods)[length(mods)] <- paste(model,names(MMods[M]),sep=".")
      # r-squared, in this case we are using the 'conditional r-squared', which includes both fixed and random effects 
      R2_mixed <- append(R2_mixed,MuMIn::r.squaredGLMM(mixed_model)[, "R2c"])
      # sigma
      sigs_mixed <- append(sigs_mixed,sigma(mixed_model))
      # Aikaike and Bayes Information Criteria
      AIC_mixed <- append(AIC_mixed,AIC(mixed_model))
      BIC_mixed <- append(BIC_mixed,BIC(mixed_model))
      ## Root mean squared error
      RMSE_mixed <- append(RMSE_mixed ,sqrt(mean(resid(mixed_model)^2)))
      ## The standardized RMSE, again against the mean of the response variable
      RMSE_mixed_std = append(RMSE_mixed_std,sqrt(mean(resid(mixed_model)^2))/mean(data_clean[, paste0("log", responseVar)],na.rm=TRUE))
      ## The mean of the cross validated RMSE
      RMSE_CVmean_mixed <- append(RMSE_CVmean_mixed , mean(sqrt(cv$details$criterion),na.rm=TRUE))
      ## The standard deviation of the cross validated RMSE
      RMSE_CVsd_mixed <- append(RMSE_CVsd_mixed , sd(sqrt(cv$details$criterion),na.rm=TRUE))
      #  variance inflation factor
      if (k>1){VIF_mixed <- car::vif(mixed_model)}else{VIF_mixed <-NA}
      ##  add the coefficients together
      coefs_mixed <- append(coefs_mixed,c(fixef(mixed_model,add.dropped=TRUE),VIF_mixed))
      #  get the intraclass correlation coefficient
      ICC <- append(ICC,icc(mixed_model)[1])
      } else {
      ####  if the model could not run, set its metrics and coefficients to NA
      predsMM <- cbind(predsMM,rep(NA,nrow(predsMM)))
      R2_mixed <- append(R2_mixed,NA)
      sigs_mixed <- append(sigs_mixed,NA)
      AIC_mixed <- append(AIC_mixed,NA)
      BIC_mixed <- append(BIC_mixed,NA)
      RMSE_mixed <- append(RMSE_mixed,NA)
      RMSE_mixed_std <- append(RMSE_mixed_std,NA)
      RMSE_CVmean_mixed <- append(RMSE_CVmean_mixed,NA)
      RMSE_CVsd_mixed <- append(RMSE_CVsd_mixed,NA)
      ICC <- append(ICC,NA)
      VIF_mixed <- append(VIF_mixed,NA)
      coefs_mixed[[M]] <- setNames(rep(NA, length(log_preds) + 1),
                                   c("(Intercept)", log_preds))
      }
  }
  ### with both the fixed and mixed effects models fit for the same set of predictors, we add their collected information to a dataset where
  ##  they can be more easily compared. 
  result_row <- tibble(
    VarGroup = varGroup,
    Model = paste(preds, collapse = ", "),
    NumPredictors = length(preds),
    Rsq_Fixed = R2_fixed,
    Sig_Fixed = sig_fixed,
    AIC_Fixed = AIC_fixed,
    BIC_Fixed = BIC_fixed,
    RMSE_Fixed = RMSE_fixed,
    RMSE_Fixed_Std = RMSE_fixed_std,
    RMSE_CVmean_Fixed = RMSE_CVmean_fixed,
    RMSE_CVsd_Fixed = RMSE_CVsd_fixed,
    ) |> mutate(MixedEffects=ME) |>
    bind_cols(as_tibble_row(setNames(as.list(c(R2_mixed,sigs_mixed,AIC_mixed,BIC_mixed,RMSE_mixed,RMSE_mixed_std,RMSE_CVmean_mixed,RMSE_CVsd_mixed,ICC)),
                                     paste0(rep(c("Rsq_Mixed","Sig_Mixed","AIC_Mixed","BIC_Mixed","RMSE_Mixed","RMSE_Mixed_Std","RMSE_CVmean_Mixed","RMSE_CVsd_Mixed","ICC_Mixed"),each=2),
                                            c("Int","IntSlope")))))
  coef_row <- tibble(
    VarGroup = varGroup,
    MixedEffects=ME,
    Model = paste(preds, collapse = ", "),
    NumPredictors = length(preds)) |>
    bind_cols(fixed_cols)|>
    bind_cols(as_tibble_row(setNames(as.list(coefs_mixed),
                                     paste0(nmsm, "_Mixed",rep(c("Int","IntSlope"),each=length(nmsm))))))
  results[[length(results) + 1]] <- result_row
  coefs[[length(coefs) + 1]] <- coef_row
}

```

The result of the above is a dataframe containing relevant metrics and model coefficents for all of the models within a given 'family', that is all of the models that use the same set of predictors but with different combinations of mixed effects. This is done for all of the families, and the results are combined together into a master data frame containing all of the information from all of the models. 

```{r ME-mods,echo=FALSE,message=FALSE,warning=FALSE}
results <- list()
coefs <- list()
MMods <- mods$mods[grep("7.M",names(mods$mods))]
MMods <- MMods[1:6]
ME <- sapply(strsplit(names(MMods),"_"),"[[",2)
for (me in unique(ME)){
  MMods2 <- MMods[ME==me]
  R2_mixed <- sigs_mixed <- AIC_mixed <-BIC_mixed<-RMSE_mixed<-coefs_mixed <- ICC <- VIF_mixed <- list()
  for (M in 1:length(MMods2)){
    mixed_model <-MMods[[M]]
    R2_mixed <- append(R2_mixed,MuMIn::r.squaredGLMM(mixed_model)[, "R2c"])
    sigs_mixed <- append(sigs_mixed,sigma(mixed_model))
    AIC_mixed <- append(AIC_mixed,AIC(mixed_model))
    BIC_mixed <- append(BIC_mixed,BIC(mixed_model))
    RMSE_mixed <- append(RMSE_mixed ,sqrt(mean(resid(mixed_model)^2)))
    VIF_mixed <- car::vif(mixed_model)
    coefs_mixed <- append(coefs_mixed,c(fixef(mixed_model,add.dropped=TRUE),VIF_mixed))
    ICC <- append(ICC,icc(mixed_model)[1])
  }
  result_row <- tibble(
  VarGroup = 6,
  Model = paste(c("Height.m","DBH.cm"), collapse = ", "),
  NumPredictors = 2,
  Rsq_Fixed = R2_fixed,
  Sig_Fixed = sig_fixed,
  AIC_Fixed = AIC_fixed,
  BIC_Fixed = BIC_fixed,
  RMSE_Fixed = RMSE_fixed,
  ) |> mutate(MixedEffects=me) |>
  bind_cols(as_tibble_row(setNames(as.list(c(R2_mixed,sigs_mixed,AIC_mixed,BIC_mixed,RMSE_mixed,ICC)),
                                 paste0(rep(c("Rsq_Mixed","Sig_Mixed","AIC_Mixed","BIC_Mixed","RMSE_Mixed","ICC_Mixed"),each=2),
                                          c("Int","IntSlope")))))
  coef_row <- tibble(
  VarGroup = 6,
  MixedEffects=ME[M],
  Model =paste(c("Height.m","DBH.cm"), collapse = ", "),
  NumPredictors = 2) |>
  bind_cols(fixed_cols)|>
  bind_cols(as_tibble_row(setNames(as.list(coefs_mixed),
                                    paste0(nmsf, "_Mixed",rep(c("Int","IntSlope"),each=length(nmsf))))))
  results[[length(results) + 1]] <- result_row
  coefs[[length(coefs) + 1]] <- coef_row
}
do.call(rbind,results)
do.call(rbind,coefs)
```

We have now collected all relevant models, with varying combinations of predictor variables and mixed effects. The 'mods' object contains the actual models as well as tables of their coefficients, various model performance metrics, and predictions from the models. 
We will now sort through, filter, and rank the models based on these performance metrics. Remember that while performance metrics are best scrutinized on models in which variables were scaled, model coefficients only make sense for unscaled variables. 
Focusing first on the model performance metrics, with unscaled data, we first arrange and rank the models based on their performance.
```{r results,message=FALSE,warning=FALSE}
##  take the results and round the values for aesthetics 
results <- mods_scaled$results%>%
  mutate(across(6:26,~round(.x,5)))
##  pivot the table and categorize metrics and models
results_long <- results %>%
  tidyr::pivot_longer(Rsq_Fixed:ICC_MixedIntSlope,names_sep = "_",names_to =c("Metric","Effects")) %>%
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) %>%
  distinct(across(1:7),.keep_all = TRUE)
##  rank the models based on their metrics
results_ranked <- results_long
results_ranked <- rbind(results_ranked%>% filter(Effects!="Fixed"),
                        results_ranked %>% filter(Effects=="Fixed")%>%
                          distinct(Model,Metric,.keep_all = TRUE)%>%
                          mutate(MixedEffects=NA))%>%
  ###  for each metric, rank the values from each model.
  ###  If R-squared, we want the highest value but all other values are optimized at the minimum
  group_by(Metric)%>%
  ###  first across all models 
  mutate(globalrank=if_else(Metric %in% c("Rsq"),rank(-value),rank(value)))%>%
  ungroup()%>%
  group_by(Metric,VarGroup)%>%
  ###  then the same within model "families"
  mutate(familyrank=if_else(Metric %in% c("Rsq"),rank(-value),rank(value)))%>%
  ungroup()%>%
  ##  now pivot back to wider with the rankings
  tidyr::pivot_wider(names_from =c("Metric"),values_from = c("value","globalrank","familyrank")) %>%
  rowwise()%>%
  ##  compute the mean of the ranks for each model family
  mutate(globalranks_mean=mean(c(globalrank_Rsq,globalrank_AIC,globalrank_BIC,globalrank_RMSE,globalrank_Sig)),
         globalranks_mean=if_else((is.na(value_ICC)|value_ICC<.1)&Effects!="Fixed",NA, globalranks_mean))%>%
  ungroup()%>%
  ###  now create a master global rank based on the means
  mutate(globalrank=rank(globalranks_mean)) %>%
  group_by(VarGroup)%>%
  ##  same for the family ranks
  mutate(familyrank=rank(globalranks_mean))%>%
  ungroup()%>%
  relocate(globalrank,familyrank,.after=Effects)%>%
  arrange(globalrank)
##  filter the ranked models. Here, we take the best model within a 'family', which is a group of models sharing the same predictors. 
results_rank_filtered <- results_ranked %>%
  tidyr::separate(Model,sep=",",into=c("Var1","Var2","Var3","Var4"),remove=FALSE)%>%
  mutate(across(c(Var1:Var4),~gsub(" ","",.x)),
         across(c(Var1:Var4),~if_else(.x=="DBH",1,if_else(.x=="Height",2,if_else(.x=="CanopyDiameter",3,4)))))%>%
  rowwise()%>%
  filter(!is.unsorted(c(Var1,Var2,Var3,Var4),na.rm=TRUE))%>%
  arrange(globalrank) %>%
  ###  Here is where we keep the highest ranking model for each family
  distinct(VarGroup,MixedEffects,Effects,.keep_all = TRUE)%>%
  relocate(ModelN) %>%
  select(-c(Var1:Var4))
##  attach the VIF values from the coefficients table, this time sourced from the unscaled data
coef <- mods$coefs |>
  mutate(across(6:32,~round(.x,5)))|>
  rowwise()|>
  mutate(maxVIF_Fixed=max(c_across(c(VIF.var1_Fixed,VIF.var2_Fixed,VIF.var3_Fixed,VIF.var4_Fixed)),na.rm=TRUE),
         maxVIF_MixedInt=max(c_across(c(VIF.var1_MixedInt,VIF.var2_MixedInt,VIF.var3_MixedInt,VIF.var4_MixedInt)),na.rm=TRUE),
         maxVIF_MixedIntSlope=max(c_across(c(VIF.var1_MixedIntSlope,VIF.var2_MixedIntSlope,VIF.var3_MixedIntSlope,VIF.var4_MixedIntSlope)),na.rm=TRUE))
coef_long <- coef %>%
  tidyr::pivot_longer('(Intercept)_Fixed':maxVIF_MixedIntSlope,names_sep = "_",names_to =c("Coefficient","Effects")) %>%
  mutate(MixedEffects=if_else(Effects=="Fixed",NA,MixedEffects),
         ModelName = if_else(is.na(MixedEffects),paste0(ModelN,".",Effects),paste0(ModelN,".",Effects,"_",MixedEffects))) %>%
  distinct(across(1:7),.keep_all = TRUE)
coef_wide <- coef_long %>%
  tidyr::pivot_wider(names_from=Coefficient)%>%
  relocate(ModelN,Effects,.before=MixedEffects)
results_rank_filtered <- results_rank_filtered %>% 
  left_join(coef_wide%>%select(ModelN,MixedEffects,Effects,maxVIF),by=join_by(ModelN==ModelN,MixedEffects==MixedEffects,Effects==Effects))%>%
  relocate(maxVIF,.after=value_ICC)
```

We now have a dataset (results_rank_filtered) with the top performing model from each family of predictor variables, ranked also by the performance against each other. Remember, however, that 
these rankings are somewhat arbitrary and based primarily upon our ability to logistically sort a large number of models. Different rankings could be applied for different objectives. Still, 
this gives us a general idea of which models are performing 'best'. We are left with around 250 models, which is still a considerable number and primarily  

We now want to test the assumptions of each model. Again, we are looking for a reasonable way to evaluate a large number of models. Using the convenient 'check_model()' function from the performance package, we can produce excellent visualizations for pertinent assumptions. These are stored in the 'Assumptions' folder of the repository. Upon examining the graphics, it seems many of the top models
```{r model-assumptions,message=FALSE,warning=FALSE}
######  testing assumptions on the top ranked models
checks <- lapply(unique(results_rank_filtered$ModelName),function(x){
  pdf(file = paste0("./Assumptions/",x,"-model_checks.pdf"), width = 9, height = 11)
  print(performance::check_model(mods$mods[[x]]))
  dev.off()
})
```


```{r results-vis,echo=FALSE,message=FALSE,warning=FALSE}
predsMM <-mods$predsMM
predsF <-mods$predsF
####  variation in models at each site?
gg_preds1 <- predsMM %>%
  #select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>%
  left_join(predsF |>select(ID,contains("preds_")),by=join_by(ID))%>%
  tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>%
  tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>%
  mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars),
         RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")),
         RandomEffects = gsub("^.*?\\.","",RandomEffects)) %>%
  filter(!grepl("Site",RandomVars))%>%
  ggplot(aes(x=logDBH.cm,y=logAGB.kg,col=Species))+
  geom_point(alpha=1)+
  geom_line(aes(y=value,lty=RandomEffects))+
  #geom_smooth(method="lm",formula=y~x,aes(lty=RandomEffects),se=FALSE)+
  facet_grid(Species~Site)+
  #scale_color_manual(values=c("black",RColorBrewer::brewer.pal(3,"Set1")))+
  scale_linetype_manual(values=c(1,2,4))+
  theme_bw()
gg_preds2 <- predsMM %>%
  #select(Dataset:logWoodDensity,contains("mod1_")) %>%
  cbind(predsF %>% select(contains("preds_")))%>%#rename(mod1_none_none = mod1_)) %>%
  tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>%
  tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>%
  mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars),
         RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")),
         RandomEffects = gsub("^.*?\\.","",RandomEffects)) %>%
  ggplot(aes(x=logDBH.cm,y=logAGB.kg,shape=Species))+
  #geom_point(col="darkgrey")+
  #geom_smooth(method="lm",formula=y~x,aes(lty=RandomEffects),se=FALSE)+
  facet_wrap(~Species)+
  scale_color_manual(values=c("black",RColorBrewer::brewer.pal(3,"Set1")))+
  scale_linetype_manual(values=c(1,2,4))+
  geom_line(aes(y=value,lty=RandomEffects,col=RandomVars))+
  theme_bw()

#####   Residual density plot
predsMM %>%
  select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>%
  cbind(predsF %>% select(contains("preds_")))%>%
  tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>%
  tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>%
  mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars),
         RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")),
         RandomEffects = gsub("^.*?\\.","",RandomEffects),
         Residuals = logAGB.kg-value) %>%
  ggplot(aes(x=Residuals,col=RandomVars))+
  geom_density()+
  facet_wrap(~RandomEffects)+
  theme_bw()

###  Residuals versus observations
predsMM %>%
  select(Dataset:logWoodDensity.g.cm3,contains("preds_")) %>%
  cbind(predsF %>% select(contains("preds_")))%>%
  tidyr::pivot_longer(cols=contains("preds_"),names_prefix = "preds_") %>%
  tidyr::separate(name,sep="(_)",remove=FALSE, into=c("RandomEffects","RandomVars"))%>%
  mutate(RandomVars = if_else(is.na(RandomVars), "none",RandomVars),
         RandomVars = factor(RandomVars,levels=c("Species&Site","Species","Site","none")),
         RandomEffects = gsub("^.*?\\.","",RandomEffects),
         Residuals = logAGB.kg-value) %>%
  ggplot(aes(x=value,y=Residuals,col=RandomEffects))+
  geom_point()+
  facet_wrap(~RandomVars)+
  theme_bw()
```

